---
title: "Kde in tidytracks"
author: "Andrea"
date: "2026-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load the boar data
```{r}
library(tidytracks)
wildboars_tt <- readRDS("./articles_data/wildboar_tt.rds")
wildboars_tt
```

Check the metadata
```{r}
show_meta(wildboars_tt)
```

Compute the kde
```{r}
wildboars_tt <- wildboars_tt %>% dplyr::group_by(name)
wildboars_kde <- hr_kde(wildboars_tt, levels = NULL,
                           h = "h_ref_indiv",
                           res = 194.7458,
                           bbox = list(xmin = 695842.0, ymin = 3154018.0,
                                    xmax = 704216.0, ymax = 3165702))
```

Check the output
```{r}
wildboars_kde
```

The h-values are the same as those computed with adehabitatHR

```{r}
sum(wildboars_kde$kde[[1]]$z)
sum(wildboars_kde$kde[[2]]$z)
sum(wildboars_kde$kde[[3]]$z)
sum(wildboars_kde$kde[[4]]$z)

```
These values are not the same as when we set same4all=TRUE in adehabitatHR 
(we have 2.636395e-05 there). We need to check whether we are using exactly
the same grid and resolution (we have the correct h values though).

Same grid corner, but different resolution (we have 252.9354 vs 194.7458 in adehabitatHR in the `kus` object from the other vignette).

