---
title: "Testing UDs in adehabitatHR"
output: html_document
---

Here we are running through the functions for making and comparing utilization
distributions (UDs) in the R package `adehabitatHR` (Calenge 2006) so that we can
sort out the logic of the UD functions in `tidytracks` and compare the outputs.

## load packages

```{r load_packages}

library(adehabitatHR)

```

## load data

The adehabitatHR package has function for estimating home ranges from animal
tracking data.

The functions accept an object of class `SpatialPoints` (one animal) or
`SpatialPointsDataFrame` (multiple animals).

The example boar tracking data has environmental variables in `map` and 
tracking data in `relocs`. We're interested in the latter, which are stored 
as a `SpatialPointsDataFrame`.

```{r load_data}

data(puechabonsp)
class(puechabonsp)
names(puechabonsp)

# two spatial points data frames
summary(puechabonsp)

# the tracking locations are in relocs
head(as.data.frame(puechabonsp$relocs))

# map of elevation
image(puechabonsp$map, col=grey(c(1:10)/10))
# add relocs
plot(puechabonsp$relocs, add=TRUE,
     col=as.data.frame(puechabonsp$relocs)[,1])

# extract relocs only as an object to use
relocs <- puechabonsp$relocs
class(relocs)
head(relocs)
```

## MCP

Compute 95% MCP for each animal. The output is stored as a 
`SpatialPolygonsDataFrame`.

```{r mcp}
# compute mcp using just the first column (coordinates) of relocs
mcp95 <- mcp(relocs[,1], percent=95)

# spatial polygons data frame with area calculated
mcp95

# plot polygons with points on top
plot(mcp95)
plot(relocs, add=TRUE)

```

## Kernel UD

"An animal's use of space can be described by a bivariate probability density
function, the UD, which gives the probability density to relocate the animal at
any place according to the coordinates (x, y)"

The kernel method is used to estimate the UD using the tracking data.

"The basic principle of the method is the following: a bivariate kernel function
is placed over each relocation, and the values of these functions are averaged
together."








