---
title: "reading_data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reading_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidytracks)
```

# Reading data into a `move2` object
The `tidytracks` package is designed to work with `move2` objects, which are a type of spatial data structure used for tracking animal movements. In `tidytrackss`, we start by reading a table of events (e.g., GPS locations). This table
(stored as a csv file) should contain at least the following columns:
- one column which is the `track_id` (i.e. the variable that groups events into a track)
- two columns representing the x and y coordinates (e.g. longitude and latitude)
- a date-time column (or separate date and time columns)

Additional columns of data will be stored in the events table if they have information
that is specific to each event (i.e. the values are not unique within a given track), or
they will be moved to the meta data table if they are track specific (e.g. bird_id, sex of the individual,
colony coordiates, breeding status, etc.).

# Example data
We start with a small dataset of wandering albatrosses from the Birdlife Seabird Tracking Database. Let's quickly
inspect it:
```{r}
albatross_csv <- system.file("/extdata/csv_files/dataset_example_birdlife.csv",
                             package = "tidytracks")
dplyr::glimpse(read.csv(albatross_csv))
```
