---
title: "tidytracks"
output: 
        rmarkdown::html_vignette
        # pdf_document: default
vignette: >
  %\VignetteIndexEntry{reading_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

# Tracking data representation in `tidytracks`

`tidytracks` uses tidy tables to represent movement, We refer to such a table as a **tibble of tracks** (technically, such tables are `move2`
objects from the `move2` package).

In a tibble of **tracks**, each row represents an **event**, which is a
time-stamped observation, associated with a location (and, optionally, extra 
data such as activity information).

**Events** are grouped into **tracks**. Initially a **track** might be all the 
locations from one deployment of a tracking device on an animal. However, for
central place foragers, we might want to split this track into separate trips 
away from the central place. In this case, the unit of tracking (i.e. the new
**track**) will become the trip, rather than the full deployment.

The **metadata** that describes each of these tracks is stored in a separate
table, which is linked to the event table by a `track_id_column`. This means we
do not have to repeat information (such as the **ring_number** or *sex* of an 
animal) for each **event**, but we can use the **metadata** of the **track** 
to store that
information.

# Reading data into a tibble of tracks

In `tidytracks`, we start by reading a table of events (e.g., GPS locations)
from a CSV file or as existing `data.frame` in R.

This table should contain at least the following columns:
- one column which is the `track_id` (i.e. the variable that groups events into 
 a track, e.g. `bird_id` or `logger_id`)
- two columns representing the `x` and `y` coordinates (e.g. `longitude` and 
 `latitude`)
- a `date-time` column (or separate `date` and `time` columns)

Additional columns of data will be stored in the **events** table if they have
information that is specific to each event (i.e. the values are not the same
within a given track, e.g. the temperature recorded at each location),
or they will be moved to the **metadata** table if they are
track specific (e.g. sex of the individual, colony coordinates,
breeding status, etc.).

## Example data

This vignette will use an example dataset of GPS tracks from the Antarctic shag
*Leucocarbo bransfieldensis*, a seabird acting as a central-place forager.
This dataset contains 9 tracks with nearly 4,000 locations. Note that these are
based on real data, but have been modified to show the functionality of
`tidytracks`. Please do not use them for any real analyses of shag movement!

We have a single CSV. Each
row includes information about the locations (in this case `lat` and
`lon`) and time (in this case, one `date_time` column). Locations belonging
to the same bird's track are identified by the same value in the `bird_id` column
(in this example, as we only have one track per bird, the `bird_id` is the individual 
ID number of the bird).

The CSV also contains other columns of information about
the tracks (`sex` and colony coordinates split into `colony_lat` and `colony_lon`).
These are repeated for each event, so they will be moved to the metadata table when 
we create the `tidytracks` object.

```{r read_events}
library(tidytracks)
library(dplyr)
library(ggplot2)
library(sf)

shags_csv <- system.file(
  "extdata",
  "shags_example.csv",
  package = "tidytracks"
)

shags_tt <- tt_read_data(
  events = shags_csv,
  col_track_id = "bird_id",
  col_coords = c("lon", "lat"),
  col_date_time = "date_time",
  time_zone = "UTC",
  crs = 4326)

```

Inspecting the events table, we have:

```{r inspect_events}
shags_tt
```

We can see how each row in the table is an event, in this case a time-stamped
GPS location for a given bird, with the column `date_time` providing the time
and column `geometry` defining the location. `bird_id` (the designated
`track_id_column`) is the identifier that groups a set of events into a single
track.

The rest of the information was moved to the meta data table. We can inspect 
the meta data table as follows:

```{r inspect_meta}
show_meta(shags_tt)
```

We have a `bird_id` column that links back to the events table, and then
information about the sex of the bird, as well as a column of
coordinates for the colony location. 

Note that `move2` objects don't have to be ordered by time, but for many analyses,
this is a requirement. We can order the data by time using the `tt_order_time()` function:

```{r order_by_time}
shags_tt <- shags_tt %>% tt_order_time()
```

Functions that start with `tt_` are specific to tidy tables of tracks (i.e.
`move2` objects), and they operate on the whole object.

# Plotting the data

To visualise the data with the `ggplot2` package, we can use the `geom_event_point()`
function to plot the locations of each event. It is good practice to project data 
each time we plot it. `geom_event_point()` is compatible 
with `coord_sf()` so we can add a projection directly into the `ggplot`. All other
decorations of `ggplot2` can be used; in this case, we will also adjust the aspect ratio 
within `theme()`.

```{r plot_events, fig.width=6, fig.height=4}
ggplot() +
  geom_event_point(data = shags_tt, aes(color = bird_id))+
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs')+
    theme(aspect.ratio = 1, 
      axis.text.x = element_text(angle = 45, hjust = 1))


```

We can explore what each individual is doing by plotting the locations of each
individual separately with `facet_wrap`. In this example, instead of plotting 
locations as points, we will use `geom_event_path()` to plot the
paths between them:

```{r shags_plot_paths, fig.width=6, fig.height=4}
ggplot() +
  geom_event_path(data = shags_tt, aes(color = bird_id)) +
  facet_wrap(~bird_id) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs') +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1))
```

We can see that bird `'kb_38'` moved very little (or not at all) during
the tracked period.

# A grammar of movement

As we saw above, `tt_*` functions are specific to and operate on the whole table
of tracks (aka the `move2` object), and often will return either a complete table of
tracks or different tibble (e.g. a tibble of kernels). For example,
there are functions to clean the data using
a given algorithm (e.g. `tt_clean_mcconnell`), or to split the data into
trips for central place foragers (e.g. `tt_split_trips`). We will see those
functions in action in a later section.

`event_*` functions, on the other hand, operate on events, and return a vector
of the same length as number of rows in the event table (and thus can be use in
`mutate()` operations). For example, we can use `event_speed()` to calculate the
speed of each event. Note that, technically, the speed pertains to the segment
between two events, but for simplicity we will refer to it as the speed of the
event. For each track, `event_speed()` returns the speed of each segment, padded
with a final NA so that we have the same number of rows as the original track.

So, we can add speed to our `shags_tt` object with:
```{r speed}
shags_tt <- shags_tt %>% 
  mutate(speed = event_speed(.))
```

Note the use of the `.` place-holder in the brackets of `event_speed()` to pass
the whole `move2` object to the function. This is only possible with a `magrittr` 
pipe (`%>%`), and not with the base R pipe (`|>`).

We can now visualise the distribution of speeds:
```{r plot_speed, fig.width=6, fig.height=4}
shags_tt %>%
  ggplot(aes(x = speed)) +
  geom_histogram(bins = 100)
```

Note the message about 9 rows having been dropped; those are the 9 last events
for which there is no distance. Also, when appropriate, functions in `tidytracks` return
values with units, to avoid conversion problems (in this case, meters per minute).
Functions usually have a `units` parameter that allows to change the units that
are returned (more details about working with units below).

The histogram above suggests that there might be outliers in the data,
as we seem to have a few events with very high speeds. We will see later
how we can clean up the data using McConnell's algorithm. 

Finally, there are `track_*` functions, which operate on the whole track, and will
return a vector of length equal to the number of tracks in the object.
For example, `track_duration()` returns the total duration of each track:

```{r duration}
shags_tt %>%
  track_duration()
```

If we wanted the same estimates in hours rather than days, we would simply say:
```{r duration_units}
shags_tt %>%
  track_duration(units = "hours")
```

This naming convention helps clarifying the scope of a given function, avoiding
ambiguities. 

For example, `event_flag_mcconnell()` will return a logical vector
indicating whether events are valid or whether they should be filtered out using
McConnell's algorithm, whilst `tt_clean_mcconnell()` will return tibble of tracks
with the invalid events removed. 

Similarly, `event_distance()` will return a vector of distances between events
(a vector of the same length as the rows of the event column), while
`track_distance()` will return the total (cumulative) distance for each track
(and thus a shorter vector of length equal to the number of tracks in the
object).

# Cleaning the data

We can get an overview of the speeds of each step with:
```{r summary_speed}
shags_tt %>%
  summary()
```

Some of those speeds (>15000 m/minute) seem a bit high. We can check the units
(to make sure we have them right):
```{r units}
units(shags_tt$speed)
```
Let's recalculate them in units that might be more familiar to us.

```{r units2}
shags_tt <- shags_tt %>%
  mutate(speed = set_units(speed, "km/h")) 

shags_tt %>%
  summary()
```
Hmm, 904 km/h is much faster than typical flight speed for a shag (European
shags usually fly at speeds around 50–65 km/h; see e.g. Wanless et al. 1997).

Let's plot the speed of each event:
```{r plot_speed_new, fig.width=6, fig.height=4}
ggplot() +
  geom_event_path(
    data = shags_tt,
    aes(color = speed)) +
  facet_wrap(~bird_id) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs') +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1))
```

It looks like there are a few GPS location errors here, resulting in unrealistic
speeds (we can easily see the very fast outlier is at the end of a track).

We can remove locations characterised by unlikely speeds using the McConnell's
algorithm (the same used by `trip::speedfilter()`). Max speed can be set to an
appropriate number for the species we are working with. Shags have been observed
flying at 58km/h (Yukihisa et al. 2016) we will set the maximum to 60. Note that 
we can give the `max_speed` in any units we want, and they will be automatically 
converted to the units in the object.

```{r mcconnel}
n_before <- nrow(shags_tt) 

shags_tt <- shags_tt %>%
  tt_clean_mcconnell(max_speed = as_units(60, "km/h"))

n_after <- nrow(shags_tt)

n_before - n_after
```

Note that `tt_clean_mcconnell()` is "destructive" and will remove the points
flagged by the algorithm. There is also a function `event_flag_mcconnell()`
which flags points without removing them; see its help page for more
details on how to use it.

Here we see that `r (n_before - n_after)` points have been removed. 
Let us now recompute the speeds after running the filter (again, in km/h):

```{r post_mcconnel}
shags_tt <- shags_tt %>%
  mutate(speed = event_speed(., units = "km/h"))
summary(shags_tt$speed)
```

Note again the use of '.' to pass the tibble to `event_speed()`.

We can see that we have removed some locations with unrealistic speeds. Note
that the McConnell algorithm does not simply remove steps above the threshold
speed, but uses the root mean square speed for previous/next and 2nd previous to
next point, thus allowing for individual steps that might have a speed just
above the threshold if the subsequent point is not too far away.

Note: because of this, sometimes if there are two consecutive points with 
unrealistically high speeds, the algorithm may miss one, this can be remedied by 
recalculating speeds and then running the filter again. 

# Splitting trips

Shags are central place foragers, and we now want to split the track into
trips. For this we need to define the colony location, and a buffer around it.
`tt_split_trips()` can take either a single colony point, or one nest location
per track as a point column in the metadata

In `tidytracks`, all locations should be stored as `sfc_POINT` geometries with a
defined coordinate reference system (CRS). Whilst this is done automatically for
locations events, location  information in the metadata needs to converted after reading the file:

```{r add_colony_geom_to_meta}
show_meta(shags_tt) <- show_meta(shags_tt) %>%
  mutate(colony_coord = sf_point_col(colony_lon, colony_lat, crs = 4326))
```

Note that we use `sf_point_col()` to create a column. Do not replace the metadata
with a full sf tibble, as that creates problems with certain functions.

When splitting trips and defining the parameters to do so, it is important to consider
the resolution and accuracy of your locations, and the likely movements of your birds.

```{r split_trips}
shags_tt_split <- shags_tt  %>%
  tt_split_trips(
    centre_col = "colony_coord",
    buffer_outbound = as_units(3, "km"),
    buffer_inbound = as_units(3, "km"),
    complete = TRUE)
```

If we now inspect the object, we can see that there is a new column, `trip_id`,
which is also used as the new `track_id_column` in the `move2` object:
```{r inspect_splits}
head(shags_tt_split)

unique(shags_tt_split$trip_id)
```

Let's look at the metadata:
```{r split_meta}
str(show_meta(shags_tt_split))
```
We can see that all the birds except for kb_38 and kb_40 made between 1 and 3 central 
place foraging trips. kb_38 didn't move and was therefor filtered out. 
kb_40 migrated away from the colony and did not return so was also removed.

**How the trip splitting works:**

Essentially, the `buffer_outbound` and `buffer_inbound` arguments in
`tt_split_trips` define zones around the central place (colony) that an animal
must cross to be considered as starting or ending a trip. The `complete = TRUE`
argument ensures that only trips which both leave and return to the colony
(i.e., complete round trips) are retained. As a result, any tracks that do not
leave the buffer zone, or trips that are not complete, are removed from the
dataset.

```{r plot_split, fig.width=6, fig.height=4}
ggplot() +
  geom_event_path(data = shags_tt_split, aes(color = trip_id)) +
  geom_sf(data = show_meta(shags_tt_split)$colony_coord, color = "grey20")+
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs') +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1))

```


# Summarising tracks

We can get a summary of our tracks with:

```{r summary}
shags_tt_split %>%
  track_summary_stats(centre_col = "colony_coord")
```

Note that `track_summary_stats()` will return a tibble with one row per track,
rather than the usual vector with one element per track that is produced by most
`track_*` functions. Note that these distance calculations are being done on unprojected 
data in Euclidean distance. Reprojection within the ggplot maps only applies within the map 
itself and doesn't change the tidytracks object.


## Filter trips

We will now filter the data set to leave only the females, so that we can then
perform some analysis. This information is the meta table, so we will use
`filter_by_meta()` to filter the dataset:


```{r filter_females}
shags_females <- shags_tt_split %>%
  filter_by_meta(sex == "female")
show_meta(shags_females)
```

We can now plot the trips:
```{r plot_females, fig.width=6, fig.height=4}
ggplot() +
  geom_event_path(
    data = shags_females,
    aes(color = bird_id) # can't use trip_id here because not trip splitted data
  ) +
  facet_wrap(~bird_id) +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position="none")
```

If we wanted, we could also add the track summary statistics into the metadata, 
and use this information to filter the data set, for example to only trips that 
were longer than a certain duration or distance threshold.

```{r add_summary_stats}
show_meta(shags_tt_split) <- show_meta(shags_tt_split) %>%
  left_join(track_summary_stats(shags_tt_split, centre_col = "colony_coord"), 
            by = "trip_id")

show_meta(shags_tt_split)
```

We may wish to move metadata to the events table (using `as_event_column()`)
or vice versa (using `as_meta_column()`).
For example, to count the number of points available for males and females,
we can move the sex information into the event table, and count how many events
we have for males and females:

```{r moving_sex_to_event}
shags_tt_split %>%
  as_event_column(sex) %>%
  pull(sex) %>%
  table()
```


# Home Range estimates

For this analysis, we will use the shags_females data set we just created. To 
estimate home ranges, we need to project the data to an equal area projection. 
We can use the `sf::st_transform()` function to do this:
```{r female_crs}
shags_females_proj <- shags_females %>%
  st_transform(crs = "+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs")
```

We can now estimate the home range using Kernel Density Estimation (KDE) with the
function `tt_hr_kde`, which  creates a KDE per trip, normalises them, then combines 
into one KDE per group that you specify. here we will group by by individual but you 
could instead group by e.g. sex.

The default smoothing parameter (h) in `tt_hr_kde` 
is set to "h_ref_mean". The bandwidth for the kernel density estimation can be either a 
number, or "h_ref_indiv" for using the reference bandwidth for each individual, or 
"h_ref_mean" for using the mean bandwidth for all individuals (the default). 

We want to estimate the home range for each bird, so we will group by `bird_id`:
```{r}
shags_females_kde <- shags_females_proj %>% # doesnt work on trip splitted data
  group_by(bird_id) %>% 
  tt_hr_kde(levels = c(0.5, 0.95))
```

We get a warning since one of the dataset is small, and so the 0.50 isopleth could not be computed. This is because the KDE is based on a grid, and if there are too few points, the grid will not be able to capture the distribution of the points.

`shags_females_kde` is an `sf` object, with estimates of the area (in the units of
the projection, in this case m^2) and a geometry column:
```{r}
shags_females_kde
```

We could easily change the units of the area with `set_units()`:
```{r}
shags_females_kde %>%
  mutate(area = set_units(area, "km^2")) %>%
  select(bird_id, level, area)
```

If you want to NOT do KDE per trip, you need to change the track_id field of your 
tidytracks object. For example if you already trip splitted, the unit of tt object 
is now trip, and kernels will always be created per trip and aggregated to your 
grouping. If you're then doing something else like analysis of the whole incubation 
period and you want to compare the kernels at the whole-bird level, you need to 
change your inherent tt object track_id to the bird_id instead.

We can plot the home ranges:

```{r plot_kde, fig.width=6, fig.height=4}
ggplot(shags_females_kde) +
  geom_sf(aes(fill = bird_id), alpha = 0.7) +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1))

```

We can add actual events with:
```{r plot_kde_points, fig.width=6, fig.height=4}
ggplot() +
  geom_sf(
    data = shags_females_kde,
    aes(fill = bird_id), alpha = 0.7
  ) +
  geom_event_point(
    data = shags_females_proj,
    size = 0.1
  ) +
  facet_wrap(~bird_id) +
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1))
```

To get a minimum convex polygon covering 95% of the data, we can use:
```{r compute_mcp}
shags_females_mcp <- shags_females_proj %>%
  group_by(bird_id) %>%
  tt_hr_mcp(levels = c(0.95))
```

We can plot the MCPs with:
```{r plot_mcp, fig.width=6, fig.height=4}
ggplot() +
  geom_sf(
    data = shags_females_mcp,
    aes(fill = bird_id), alpha = 0.7
  ) +
  geom_event_point(
    data = shags_females_proj,
    size = 0.1
  ) +
  facet_wrap(~bird_id)+
  theme(aspect.ratio = 1, 
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Save your data
We can save our `move2` object with the `tt_write_data()` function, which can save the events and meta data either separately or as a single CSV text file. In this example we will save it to the temporary directory.

```{r save_data}

# save to temp directory
tmp_prefix <- file.path(tempdir(), "shags_females")

tt_write_data(
  shags_females,
  tmp_prefix,
  combined = TRUE
)

```








