---
title: "tidytrack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidytrack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidytrack)
```

We use a dataset on boobies from the `track2KBA` package. The dataset contains 
the tracking data of 43 boobies. We will use this dataset to demonstrate the 
functionalities of the `tidytrack` package.

```{r echo=FALSE, results='hide'}
library(dplyr)
# remove colony information from the dataset
boobies <- track2KBA::boobies |> select(-c(lon_colony, lat_colony))

```

Let's inspect the table
```{r}
summary(boobies)
```

We start by combining the `data_gmt` and `time_gmt` columns into a single `POSIXct`,
using the `lubridate` package, and then removing the original columns. 
```{r}
library(dplyr)
library(lubridate)
boobies <- boobies |>
  mutate(date_time = ymd_hms(paste(date_gmt, time_gmt))) |>
  select(-c(date_gmt, time_gmt))
head(boobies)
```

Next, we convert the data into an `sf` object.
```{r}
library(sf)
boobies_sf <- st_as_sf(boobies, coords = c("longitude", "latitude"),
                       crs = "+proj=longlat +datum=WGS84")
summary(boobies_sf)
```

We are now ready to cast the data into a `move2` object, which we will then
use as the base of our analysis:
```{r}
library(move2)
boobies_move <- mt_as_move2(boobies_sf, track_id_column = "track_id",
                            time_column = "date_time")
```

We can now print a summary of the `move2` object:
```{r}
boobies_move
```

Each row in the table is an event, in this case a time-stamped GPS location for 
a given bird, with the column `date_time` providing the time and column `geometry`
defining the location. `track_id` (the designated `track_id_column`) is 
is the identifier that groups a set of events into a single track. Each track 
can further be given information in the form of metadata (metadata refers to
information that applies to the whole track, like the id of the bird or its sex).
We can access this information with `mt_show_meta()`:
```{r}
mt_show_meta(boobies_move)
```

In this case, the metadata tibble only includes the `track_id`.
