---
title: "tidytrack"
output: # rmarkdown::html_vignette
        pdf_document: default
vignette: >
  %\VignetteIndexEntry{tidytrack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidytrack)
```

We use a dataset on boobies from the `track2KBA` package. The dataset contains 
the tracking data of 43 boobies. We will use this dataset to demonstrate the 
functionalities of the `tidytrack` package.

```{r echo=FALSE, results='hide'}
library(dplyr)
# remove colony information from the dataset
boobies <- track2KBA::boobies |> select(-c(lon_colony, lat_colony)) |>
  mutate(track_id = as.factor(track_id))

```

Let's inspect the table
```{r}
summary(boobies)
```

We start by combining the `data_gmt` and `time_gmt` columns into a single `POSIXct`,
using the `lubridate` package, and then removing the original columns. 
```{r}
library(dplyr)
library(lubridate)
boobies <- boobies |>
  mutate(date_time = ymd_hms(paste(date_gmt, time_gmt))) |>
  select(-c(date_gmt, time_gmt))
head(boobies)
```

Next, we convert the data into an `sf` object.
```{r}
library(sf)
boobies_sf <- st_as_sf(boobies, coords = c("longitude", "latitude"),
                       crs = "+proj=longlat +datum=WGS84")
summary(boobies_sf)
```

We are now ready to cast the data into a `move2` object, which we will then
use as the base of our analysis:
```{r}
library(move2)
boobies_move <- mt_as_move2(boobies_sf, track_id_column = "track_id",
                            time_column = "date_time")
```

We can now print a summary of the `move2` object:
```{r}
boobies_move
```

Each row in the table is an event, in this case a time-stamped GPS location for 
a given bird, with the column `date_time` providing the time and column `geometry`
defining the location. `track_id` (the designated `track_id_column`) is 
is the identifier that groups a set of events into a single track. Each track 
can further be given information in the form of metadata (metadata refers to
information that applies to the whole track, like the id of the bird or its sex).
We can access this information with `mt_show_meta()`:
```{r}
mt_show_meta(boobies_move)
```

As we did not import any additional metadata information, the metadata tibble
only includes the `track_id`. We can add more information by loading a table
of metadata and joining it to the current table:

TODO

Note that `move2` objects don't have to be ordered by time. We can order the
data by time using the `mt_order()` function:
```{r}
boobies_move <- boobies_move |> mt_order_time()
```



# Plotting the data

We can plot the data using the `ggplot2` package. Let us filter out the tracks
for individuals 69302, 69303, and 69304. We could filter directly the `move2`
object as if it was a normal tibble. However, we can also use information from
the metadata to filter the tracks. We can use the `filter_by_meta()` function
to do so:

```{r}
boobies_sub <- boobies_move |> filter_by_meta(track_id %in% c("69302", "69303", "69304"))
boobies_sub
```

We can now plot the data:
```{r}
library(ggplot2)
ggplot() + geom_sf(data = boobies_sub, aes(color = track_id))
```

We can only see two individuals in the plot. Let's explore what each individual
is doing by plotting the locations of each individual separately:
```{r}
ggplot() +
  geom_sf(data = boobies_sub, aes(color = track_id)) +
  facet_wrap(~track_id)
```
We can see that '69303' moved very little (or not at all) during the period.

As a `move2` object is a derived `sf` tibble, we can simply use `coord_sf` to
plot the data in a map projection:
```{r}
ggplot() +
  geom_sf(data = boobies_sub, aes(color = track_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

Instead of plotting the individual locations, we might prefer to plot the
trajectory using `mt_track_lines()`:

```{r}
ggplot() +
  geom_sf(data = mt_track_lines(boobies_sub), aes(color = track_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

# Cleaning the data

We can get an overview of the speeds of each step with:
```{r}
boobies_sub |> mt_speed() |> summary()
```

Hmm, there is something odd here, as the maximum speed is 2044 m/minute. Incidentally,
if we just print a few speeds, we can easily check the units:
```{r}
boobies_sub[1:10,] |> mt_speed()
```

We can remove locations characterised by unlikely speeds using the McConnell's
algorithm (the same used by `trip::speedfilter()`):

```{r}
boobies_sub_cleaned <- boobies_sub |>
  mt_clean_mcconnell(max_speed = units::as_units(50,"m/min"))
```

Note that `mt_flag_mcconnell()` is "destructive" and will remove the points
flagged by the algorithm. There is also a function `event_flag_mcconnell()`
which flags points without removing them; see the advanced vignette for more
details on how to use it (TODO there is no advance vignette yet!!!).

Let us now check speeds:

```{r}
boobies_sub_cleaned |> mt_speed() |> summary()
```

We can see that we have removed extreme speeds.

Now we can plot the cleaned data:
```{r}
ggplot() +
  geom_sf(data = mt_track_lines(boobies_sub_cleaned), aes(color = track_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```
