---
title: "tidytrack"
output: # rmarkdown::html_vignette
        pdf_document: default
vignette: >
  %\VignetteIndexEntry{tidytrack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Tracking data representation in `tidytrack`

`tidytrack` used `move2` objects to represent tracking data. In a `move2`
object, each row represents an event, which is a time-stamped observation,
often associated with a location. Events are grouped into tracks. Tracks can
be all the locations from a deployment, or a subset of locations for a trip
of a central place forager. The metadata that describes each of this track is
stored in a separate table, which is linked to the event table by a 
`track_id_column`. In this way, we do not have to repeat information (such as the id of
an animal, or its sex) for each event, but we can use the metadata of the track
to characterise all events within that track. Let us inspect a sample dataset
that we will use throughout this vignette. The dataset is included in the
package, and can be loaded with the following command:

```{r setup}
library(tidytrack)
data(boobies_mt)
```

It is a dataset on boobies containing 43 tracks with just over 178k GPS
locations. Note that the data are based on XXX, but they were doctored to showcase
functionality of `tidytrack`; do not use them for any real analysis!!!.

We can start by printing a summary of the `move2` object:
```{r}
boobies_mt
```

We can see how each row in the table is an event, in this case a time-stamped GPS location for 
a given bird, with the column `date_time` providing the time and column `geometry`
defining the location. `deployment_id` (the designated `track_id_column`) is 
is the identifier that groups a set of events into a single track. 
We can access the metadata for the tracks with `mt_show_meta()`:
```{r}
mt_show_meta(boobies_mt)
```

We can see how we have a `deployment_id` column that links back to the events
table, and then information about the sex and species of the bird, as well as
a column of coordinates for the colony location (they are stored as `sfc_POINT`
geometries from the `sf` package, so that they come with crs information).

Note that `move2` objects don't have to be ordered by time. We can order the
data by time using the `mt_order()` function:
```{r}
boobies_mt <- boobies_mt |> mt_order_time()
```



# Plotting the data

We can plot the data using the `ggplot2` package. Let us filter out the tracks
for individuals 69302, 69303, and 69304. We could filter directly the `move2`
object as if it was a normal tibble. However, we can also use information from
the metadata to filter the tracks. We can use the `filter_by_meta()` function
to do so:

```{r}
boobies_sub <- boobies_mt |>
  filter_by_meta(deployment_id %in% c("69302", "69303", "69304"))
boobies_sub
```

We can now plot the data:
```{r}
library(ggplot2)
ggplot() + geom_sf(data = boobies_sub, aes(color = deployment_id))
```

We can only see two individuals in the plot. Let's explore what each individual
is doing by plotting the locations of each individual separately:
```{r}
ggplot() +
  geom_sf(data = boobies_sub, aes(color = deployment_id)) +
  facet_wrap(~deployment_id)
```
We can see that '69303' moved very little (or not at all) during the period.

As a `move2` object is a derived `sf` tibble, we can simply use `coord_sf` to
plot the data in a map projection:
```{r}
ggplot() +
  geom_sf(data = boobies_sub, aes(color = deployment_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

Instead of plotting the individual locations, we might prefer to plot the
trajectory using `mt_track_lines()`:

```{r}
ggplot() +
  geom_sf(data = mt_track_lines(boobies_sub), aes(color = deployment_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

# Cleaning the data

We can get an overview of the speeds of each step with:
```{r}
boobies_sub |> mt_speed() |> summary()
```

Hmm, there is something odd here, as the maximum speed is 2044 m/minute. Incidentally,
if we just print a few speeds, we can easily check the units:
```{r}
boobies_sub[1:10,] |> mt_speed()
```

We can remove locations characterised by unlikely speeds using the McConnell's
algorithm (the same used by `trip::speedfilter()`):

```{r}
boobies_sub_cleaned <- boobies_sub |>
  mt_clean_mcconnell(max_speed = units::as_units(50,"m/min"))
```

STOP THIS IS NOT WORKING YET!!!!

Note that `mt_flag_mcconnell()` is "destructive" and will remove the points
flagged by the algorithm. There is also a function `event_flag_mcconnell()`
which flags points without removing them; see the advanced vignette for more
details on how to use it (TODO there is no advance vignette yet!!!).

Let us now check speeds:

```{r eval=FALSE}
boobies_sub_cleaned |> mt_speed() |> summary()
```

We can see that we have removed extreme speeds.

Now we can plot the cleaned data:
```{r eval=FALSE}
ggplot() +
  geom_sf(data = mt_track_lines(boobies_sub_cleaned), aes(color = deployment_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```


# Splitting trips

Boobies are central place foragers, and we now want to split the track into
trips. For this we need to define the colony location, and a buffer around it.
`mt_split_trips()` can take either a single colony point, or one nest location
per track. 

```{r}
boobies_sub |> 
  mt_split_trips(center_col = "colony_coord",
                 buffer_outbound = as_units(1, "km"),
                 buffer_inbound = as_units(2, "km"))

```
