---
title: "tidytracks"
output: 
        rmarkdown::html_vignette
        # pdf_document: default
vignette: >
  %\VignetteIndexEntry{reading_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Tracking data representation in `tidytracks`

`tidytracks` uses tidy tables to represent movement data based on `move2`
objects.

In a tidy table of **tracks**, each row represents an **event**, which is a
time-stamped observation, associated with a location (and, optionally, extra 
data such as activity information).

**Events** are grouped into **tracks**. Initially a **track** might be all the 
locations from one deployment of a tracking device on an animal. However, for
central place foragers, we might want to split this track into separate trips 
away from the central place. In this case, the unit of tracking (i.e. the new
**track**) will become the trip, rather than the full deployment.

The **metadata** that describes each of these tracks is stored in a separate
table, which is linked to the event table by a `track_id_column`. This means we
do not have to repeat information (such as the *id* or *sex* of an animal) for
each **event**, but we can use the **metadata** of the **track** to store that
information.

This vignette will use an example dataset of GPS tracks from the Antarctic shag
*Leucocarbo bransfieldensis*, a seabird acting as a central-place forager.
This dataset contains 9 tracks with nearly 8,000 locations. Note that these are
based on real data, but have been modified to show the functionality of
`tidytracks`. Please do not use them for any real analyses of shag movement!

# Reading data into a `move2` object

In `tidytracks`, we start by reading a table of events (e.g., GPS locations)
from a CSV file or as existing `data.frame` in R.

This table should contain at least the following columns:
- one column which is the `track_id` (i.e. the variable that groups events into 
 a track)
- two columns representing the `x` and `y` coordinates (e.g. `longitude` and 
 `latitude`)
- a `date-time` column (or separate `date` and `time` columns)

Additional columns of data will be stored in the **events** table if they have
information that is specific to each event (i.e. the values are not unique
within a given track, e.g. the temperature recorded at each location),
or they will be moved to the **metadata** table if they are
track specific (e.g. track_id, sex of the individual, colony coordinates,
breeding status, etc.).

## Example data

We start with a small data set of Antarctic Shags. We have a single CSV. Each
row must include information about the locations (in this case `latitude` and
`longitude`) and time (in this case, separated into two columns, `date_gmt` and `time_gmt`). Locations belonging
to the same track are identified by the same value in the `track_id` column
(in this example, as we only have one track per bird, the `track_id` is the ring
number of the bird). 
The CSV also contains other columns of information about the tracks (e.g. `sex`, `colony`, `year`), but these are repeated for each event, so they will be moved to the metadata table when we create the tidytracks object.


```{r read_events}
library(tidytracks)
library(dplyr)
library(ggplot2)

# load data called "shags" from package
data("shags_tt", package = "tidytracks")
```

## We want to revert track_id to bird_id
## Use tt_read_data to read the data instead of using data()

Inspecting the events table, we have:

```{r inspect_events}
shags
```

We can see how each row in the table is an event, in this case a time-stamped
GPS location for a given bird, with the column `date_time` providing the time
and column `geometry` defining the location. `track_id` (the designated
`track_id_column`) is the identifier that groups a set of events into a single
track.

The rest of the information was moved to the meta data table. We can inspect 
the meta data table as follows:

```{r inspect_meta}
show_meta(shags)
```

We have a `track_id` column that links back to the events table, and then
information about the sex and species of the bird, as well as a column of
coordinates for the colony location. In `tidytracks`, all locations
should be stored as `sfc_POINT` geometries with a defined coordinate reference system (CRS).
Whilst this is done automatically for locations events, location information in the
metadata needs to converted after reading the file:

```{r add_colony_geom_to_meta}
# add geometry column (colony location point) to metadata
show_meta(shags) <- show_meta(shags) %>%
  mutate(colony_coord = sf_point_col(x = colony_lon, y = colony_lat,
                                     crs = 4326))
show_meta(shags)
```

Note that `move2` objects don't have to be ordered by time, but for many analysis,
this is a requirement. We can order the
data by time using the `tt_order_time()` function:
```{r order_by_time}
shags <- shags %>% tt_order_time()
```

Functions that start with `tt_` are specific to tidy tables of tracks (i.e.
`move2` objects), and they operate on the whole object.

# Plotting the data

To plot the data with the `ggplot2` package, we can use the `geom_event_point()` function to plot the locations of each event: 

```{r plot_data}
library(ggplot2)
ggplot() +
  geom_event_point(data = shags, aes(color = track_id))
```

We can explore what each individual is doing by plotting the locations of each
individual separately with `facet_wrap`. In this example, instead of plotting individual
locations as points, we can also use `geom_event_path()` to plot the paths between the locations:

```{r}
ggplot() +
  geom_event_path(data = shags, aes(color = track_id)) +
  facet_wrap(~track_id)
```

We can see that bird `'KB_23_42962_38'` moved very little (or not at all) during
the tracked period.

`geom_event_point()` and `geom_event_path()` are compatible with `coord_sf()`
to project the map:

# can we add this point into one of the prev maps?

```{r plot_projection}
ggplot() +
  geom_event_point(data = shags, aes(color = track_id)) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs')
```


# A grammar of movement

As we saw above, `tt_*` functions are specific to and operate on the whole table
of tracks (aka the `move2` object), and often will return a complete table of
tracks. There are functions to clean the data using
different algorithms (e.g. `tt_clean_mcconnell`), or to split the data into
trips for central place foragers (e.g. `tt_split_trips`). We will see those
functions in action in a later section.

`event_*` functions, on the other hand, operate on events, and return a vector
of the same length as number of rows in the event table (and thus can be use in
`mutate()` operations). For example, we can use `event_speed()` to calculate the
speed of each event. Note that, technically, the speed pertains to the segment
between two events, but for simplicity we will refer to it as the speed of the
event. For each track, `event_speed()` returns the speed of each segment, padded
with a final NA so that we have the same number of rows as the original track.

So, we can add speed to our `shags` object with:
```{r}
shags <- shags %>% 
  dplyr::mutate(speed = event_speed(.))
```

Note the use of the `.` place-holder in the brackets of `event_speed()` to pass
the whole `move2` object to the function. This is only possible with a `magrittr` pipe
(`%>%`), and not with the base R pipe (`|>`).

We can now visualise the distribution of speeds:
```{r}
shags %>%
  ggplot(aes(x = speed)) +
  geom_histogram(bins = 100)
```

Note the message about 9 rows having been dropped; those are the 9 last events
for which there is no distance. Also, functions in `tidytracks` tend to return
values with units, to avoid conversion problems (in this case, meters per minute).
Functions usually have a `units` parameter that allows to change the units that
are returned (more details about working with units below).

The histogram above is suggests that there might be outliers in the data,
as we seem to have a few events with very high speeds. We will see later
how we can clean up the data using McConnell's algorithm. 

Finally, there are `track_*` functions, which operate on the whole track, and will
generally return a vector of length equal to the number of tracks in the object.
For example, `track_duration()` returns the total duration of each track:

```{r}
shags %>%
  track_duration()
```

If we wanted the same estimates in hours rather than days, we would simply say:
```{r}
shags %>%
  track_duration(units = "hours")
```

This naming convention helps clarifying the scope of a given function, avoiding
ambiguities. 

For example, `event_flag_mcconnell()` will return a logical vector
indicating whether events are valid or whether they should be filtered out using
McConnell's algorithm, whilst `tt_clean_mcconnell()` will return a `move2`
object with the invalid events removed. 

Similarly, `event_distance()` will return a vector of distances between events
(a vector of the same length as the rows of the event column), while
`track_distance()` will return the total (cumulative) distance for each track
(and thus a shorter vector of length equal to the number of tracks in the
object).

<br>

***

# Cleaning the data

We can get an overview of the speeds of each step with:
```{r}
shags %>%
  summary()
```

Some of those speeds (>15000 m/minute) seem a bit high. We can check the units
(to make sure we have them right):
```{r}
units(shags$speed)
```
Let's recalculate them in units that might be more familiar to us.

```{r}
shags <- shags %>%
  mutate(speed = set_units(speed, "km/h")) 

shags %>%
  summary()
```
Hmm, 900 km/h is much faster than typical flight speeds for a shag (European
shags usually fly at speeds around 50–65 km/h; see e.g. Wanless et al. 1997).

Let's plot the speed of each event:
```{r}
ggplot() +
  geom_event_path(
    data = shags,
    aes(color = speed)
  ) +
  facet_wrap(~track_id)
```

It looks like there are a few GPS location errors here, resulting in unrealistic
speeds.

We can remove locations characterised by unlikely speeds using the McConnell's
algorithm (the same used by `trip::speedfilter()`). Max speed can be set to an
appropriate number for the species we are working with. Shags have been observed
flying at 60km/h so we will set the maximum to 80. Note that we can give the
`max_speed` in any units we want, and they will be automatically converted to
the units in the object.

```{r}
nrow(shags) # check number of events before filter
# run the speed filter
shags <- shags %>%
  tt_clean_mcconnell(max_speed = set_units(80, "km/h"))
nrow(shags) # check number of events after the filter
```

Note that `tt_clean_mcconnell()` is "destructive" and will remove the points
flagged by the algorithm. There is also a function `event_flag_mcconnell()`
which flags points without removing them; see the advanced vignette for more
details on how to use it (TODO there is no advance vignette yet!!!).

Let us now recompute the speeds after running the filter (again, in km/h):

```{r}
shags <- shags %>%
  mutate(speed = event_speed(., units = "km/h"))
summary(shags)
```

## remove/move the points that are still creating outliers after the filter (beginning or end of the track)?

We still have some locations with unrealistic speeds, so let's run the 
speed filter again.

TODO we still have unrealistic speeds - need to check where they're coming from 
and how to explain/fix this.

```{r}
nrow(shags) # check number of events before filter
# run the speed filter
shags <- shags %>%
  tt_clean_mcconnell(max_speed = set_units(80, "km/h"))
nrow(shags) # check number of events after the filter#

# recompute speeds and check summary
shags <- shags %>%
  dplyr::mutate(speed = event_speed(., units = "km/h"))
summary(shags)
```
TODO- removing end speeds, did lizzie update this? 

We can see that we have removed some locations with unrealistic speeds. Note
that the McConnell algorithm does not simply remove steps above the threshold
speed, but uses the root mean square speed for previous/next and 2nd previous to
next point, thus allowing for individual steps that might have a speed just
above the threshold if the subsequent point is not too far away.

# Splitting trips

Shags are central place foragers, and we now want to split the track into
trips. For this we need to define the colony location, and a buffer around it.
`tt_split_trips()` can take either a single colony point, or one nest location
per track (as a column in the metadata). 

```{r}
shags <- shags %>%
  tt_split_trips(
    centre_col = "colony_coord",
    buffer_outbound = as_units(3, "km"),
    buffer_inbound = as_units(3, "km"),
    complete = TRUE # keep only complete trips
  )
```

If we now inspect the object, we can see that there is a new column, `trip_id`,
which is also used as the new `track_id_column` in the `move2` object:
```{r}
shags
```

Let's look at the metadata:
```{r}
str(show_meta(shags))
```

TODO update this paragraph now that we're keeping all birds in: We can see that
only individual KB_23_42778_45 performed proper central place trips, and
specifically it made 3 trips. KB_23_44224_40 migrated away from the study area.
KB_23_42962_38 never moved and was therefore filtered out.

**How the trip splitting works:**

Essentially, the `buffer_outbound` and `buffer_inbound` arguments in
`tt_split_trips` define zones around the central place (colony) that an animal
must cross to be considered as starting or ending a trip. The `complete = TRUE`
argument ensures that only trips which both leave and return to the colony
(i.e., complete round trips) are retained. As a result, any tracks that do not
leave the buffer zone, or trips that are not complete, are removed from the
dataset.

```{r}
ggplot() +
  geom_sf(data = track_lines(shags), aes(color = trip_id)) +
  geom_sf(data = show_meta(shags)$colony_coord, color = "grey20")
```

## Truncate deployments so that they don't have more than 2/3 trips each.



<br>

***

# Summarising tracks

We can get a summary of our tracks with:

```{r}
shags %>%
  track_summary_stats(centre_col = "colony_coord")
```

Note that `track_summary_stats()` will return a tibble with one row per track,
rather than the usual vector with one element per track that is produced by most
`track_*` functions.

<br>

## Filter trips

We will now filter the dataset to leave only the females, so that we can then
perform some analysis. This information is the meta table, so we will use
`filter_by_meta()` to filter the dataset:


If we wanted, we could use this information to filter the dataset, for example
to only trips that were longer than a certain duration or distance threshold.

It is better to add the trip stats to the metadata table and use
`filter_by_meta()` or just get the `trip_id`s that meet the criteria and filter
directly on the `move2` object?

TODO add this code once we've enabled updating the metadata table.

```{r}
shags_females <- shags %>%
  filter_by_meta(sex == "Female") # we should make all strings not capital
show_meta(shags_females)
```

We can now plot the trips:
```{r}
ggplot() +
  geom_track_path(
    data = shags_females,
    aes(color = trip_id)
  ) +
  facet_wrap(~track_id) + theme(legend.position="none")

```


***

# Home Range estimates


To estimate home ranges, we need to project the data to an equal area
projection. We can use the `sf::st_transform()` function to do this:
```{r}
shags_females_proj <- shags_females %>%
  sf::st_transform(crs = "+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs")
```

We can now estimate the home range using Kernel Density Estimation (KDE). The 
default smoothing parameter (h) in the `tt_hr_kde` function is set to
"h_ref_mean". The bandwidth for the kernel density estimation can be either a 
number, or "h_ref_indiv" for using the reference bandwidth
for each individual, or "h_ref_mean" for using the mean bandwidth for all
individuals (the default). 

We want to estimate the home range for each bird, so we will group by `track_id`:
```{r}
shags_females_kde <- shags_females_proj %>%
  group_by(track_id) %>%
  tt_hr_kde(levels = c(0.5, 0.95))
```

`shags_females_kde` is an `sf` object, with estimates of the area (in the units of
the projection, in this case m^2) and a geometry column:
```{r}
shags_females_kde
```

We could easily change the units of the area with `set_units()`:
```{r}
shags_females_kde %>%
  mutate(area = set_units(area, "km^2")) %>%
  select (track_id, level, area)
```


We can plot the home ranges:
```{r}
ggplot(shags_females_kde) +
  geom_sf(aes(fill = track_id), alpha = 0.7)
```

We can add actual events with:
```{r}
ggplot() +
  geom_sf(
    data = shags_females_kde,
    aes(fill = track_id), alpha = 0.7
  ) +
  geom_event_point(
    data = shags_females_proj,
    size = 0.1
  ) +
  facet_wrap(~track_id)
```

To get a minimum convex polygon covering 95% of the data, we can use:
```{r}
shags_females_mcp <- shags_females_proj %>%
  group_by(track_id) %>%
  tt_hr_mcp(levels = c(0.95))
```

We can plot the MCPs with:
```{r}
ggplot() +
  geom_sf(
    data = shags_females_mcp,
    aes(fill = track_id), alpha = 0.7
  ) +
  geom_event_point(
    data = shags_females_proj,
    size = 0.1
  ) +
  facet_wrap(~track_id)
```

## Save your data
We can save our `move2` object with the `tt_write_data()` function, which can save the events and meta data either separately or as a single CSV text file.

##TODO eample of code (saving to the temp directory)
tempfile(fileext = ".csv")