---
title: "reading_data into tidytracks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reading_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidytracks)
```

# Tracking data representation in `tidytracks`

`tidytracks` uses tidy tables of tracks to represent tracking data (`move2` objects: which are a type of spatial data structure used for tracking animal movements). In a tidy table of tracks, each row represents an event, which is a time-stamped observation,
often associated with a location. Events are grouped into tracks. Tracks can
be all the locations from a deployment, or a subset of locations for a trip
of a central place forager. The metadata that describes each of these tracks is
stored in a separate table, which is linked to the event table by a 
`track_id_column`. In this way, we do not have to repeat information (such as the id of
an animal, or its sex) for each event, but we can use the metadata of the track
to characterise all events within that track. Let us inspect a sample dataset
that we will use throughout this vignette.  We can use the move2 object created in the previous section on reading in data.


# Reading data into a `move2` object
In `tidytrackss`, we start by reading a table of events (e.g., GPS locations). This table
(stored as a csv file) should contain at least the following columns:
- one column which is the `track_id` (i.e. the variable that groups events into a track)
- two columns representing the x and y coordinates (e.g. longitude and latitude)
- a date-time column (or separate date and time columns)

Additional columns of data will be stored in the events table if they have information
that is specific to each event (i.e. the values are not unique within a given track), or
they will be moved to the meta data table if they are track specific (e.g. bird_id, sex of the individual, colony coordiates, breeding status, etc.).

# Example data
We start with a small dataset of Antarctic Shags. Let's quickly
inspect it:
```{r}
shag_csv <- system.file("extdata/shag_tidytrack_sample_subset.csv",
                             package = "tidytracks")
dplyr::glimpse(read.csv(shag_csv))
```
Let us now read in the data using `tt_read_data()`. This function reads a csv file containing event data and converts it into a `move2` object.
```{r}
shags_mt <- tt_read_data(shag_csv,
                              col_track_id = "bird_id",
                              col_coords = c("longitude", "latitude"),
                              col_date_time = c("date_gmt", "time_gmt"))
#data_time still not reading in correctly
```


Inspecting the events table, we have:
```{r}
shags_mt
```

We can see that only date and locations were retained as event specific variables, while the rest of the information was moved to the meta data table. We can inspect the meta data table as follows:
```{r}
dplyr::glimpse(show_meta(shags_mt))
```

It is a dataset on antarctic shags containing 6 tracks with 7,798 GPS -->
locations. Note that the data are based on XXX, but they were doctored to showcase -->
functionality of `tidytracks`; do not use them for any real analysis!!!. --> 

We can start by printing a summary of object:
```{r}
summary(shags_mt)
```
We can see how each row in the table is an event, in this case a time-stamped GPS location for 
a given bird, with the column `date_time` providing the time and column `geometry`
defining the location. `bird_id` (the designated `track_id_column`) is 
is the identifier that groups a set of events into a single track.  #has multiple datetime colums-do we want this??

We can access the metadata for the tracks with `show_meta()`:
```{r}
head(show_meta(shags_mt))
```

We can see how we have a `bird_id` column that links back to the events
table, and then information about the sex and species of the bird, as well as
a column of coordinates for the colony location (they are stored as `sfc_POINT`
geometries from the `sf` package, so that they come with crs information).

Note that `move2` objects don't have to be ordered by time. We can order the
data by time using the `tt_order_time()` function:
```{r}
shags_mt <- shags_mt %>% tt_order_time()
```

Functions that start with `tt_` are specific to tidy tables of tracks
(i.e. `move2` objects), and they
operate on the whole object.

# Plotting the data

We can plot the data using the `ggplot2` package. Let us filter out the tracks
for individuals 69302, 69303, and 69304. We could filter directly the `move2`
object as if it was a normal tibble. However, we can also use information from
the metadata to filter the tracks. We can use the `filter_by_meta()` function
to do so:

```{r}
boobies_sub <- boobies_mt %>%
  filter_by_meta(bird_id %in% c("69302", "69303", "69304"))
boobies_sub
```

We can now plot the data:
```{r}
library(ggplot2)
ggplot() +
  geom_event_point(data = boobies_sub, aes(color = bird_id))


```

We can only see two individuals in the plot. Let's explore what each individual
is doing by plotting the locations of each individual separately:
```{r}
ggplot() +
  geom_event_point(data = boobies_sub, aes(color = bird_id)) +
  facet_wrap(~bird_id)

```

```{r}
ggplot() +
  geom_event_path(data = boobies_sub) +
  facet_wrap(~bird_id)
```


We can see that '69303' moved very little (or not at all) during the period.

As a `move2` object is a derived `sf` tibble, we can simply use `coord_sf` to
plot the data in a map projection:
```{r}
ggplot() +
  geom_event_point(data = boobies_sub, aes(color = bird_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

Instead of plotting the individual locations, we might prefer to plot the
trajectory as a path:

```{r}
ggplot() +
  geom_track_path(data = boobies_sub, aes(color = bird_id)) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

# A grammar of movement

As we saw above, `tt_*` functions are specific operate on the whole object, and
often will return a complete table of tracks (aka a `move2` object).
There are functions to clean the data using different algorithms
(e.g. `tt_clean_mcconnell`), or to split the data into trips for central place
foragers (e.g. `tt_split_trips`). We will see those functions in action in a later
section.

`event_*` functions, on the other hand, operate on events, and return a vector
of the same length as number of rows in the event table (and thus can be use in `mutate()`
operations). For example, we can use `event_speed()` to calculate the speed of 
each event. Note that, technically, the speed pertains to the segment between
two events, but for simplicity we will refer to it as the speed of the event.
For each track, `event_speed()` returns the speed of each segment, padded with a
final NA so that we have the same number of rows as the original track.

So, we can add speed to our `boobies_sub` object with:
```{r}
library(dplyr)
boobies_sub <- boobies_sub %>% mutate(speed = event_speed(.))
```

Note the use of the . place-holder in the brackets of `event_speed()` to pass
the `move2` object to the function. This is only possible with a `magrittr` pipe
(`%>%`), and not with the base R pipe (`|>`).

We can now visualise the distribution of speeds:
```{r}
boobies_sub %>%
  ggplot(aes(x = speed)) +
  geom_histogram(bins = 100)
```

Note the message about 3 rows having been dropped; those are the 3 last events
for which there is no distance. Also, functions in `tidytracks` tend to return
values with units, to avoid conversion problems (in this case, meters per minute).
Functions usually have a `units` parameter that allows to change the units that
are returned (more details about working with units below).

The histogram above is suggests that there might be outliers in the data,
as we seem to have a few events with very high speeds. We will see later
how we can clean up the data using McConnell's algorithm.

Finally, there are `track_*` functions, which operate on the whole track, and will
generally return a vector of length equal to the number of tracks in the object.
For example, `track_duration()` returns the total duration of each track:

```{r}
boobies_sub %>%
  track_duration()
```

If we wanted the same estimates in hours rather than days, we would simply say:
```{r}
boobies_sub %>%
  track_duration(units = "hours")
```

This naming convention helps clarifying the scope of a given function,
avoiding ambiguities. For example, `event_flag_mcconnell()` will return a logical vector
indicating whether events are valid or whether they should be filtered out
using McConnell's algorithm, whilst `tt_clean_mcconnell()` will return a
`move2` object with the invalid events removed. Similarly, `event_distance()` 
will return a vector of distances between events (a vector of the same length
as the rows of the event column), while `track_distance()` will
return the total (cumulative) distance for each track (and thus a shorter
vector of length equal to the number of tracks in the object).

# Cleaning the data

We can get an overview of the speeds of each step with:
```{r}
boobies_sub %>%
  summary()
```

Some of those speeds (>2000 m/minute) seem a bit high. We can check the units (to make sure
we have them right):
```{r}
units(boobies_sub$speed)
```


Let's rephrase them in units that might be more familiar to us.

```{r}
boobies_sub %>%
  mutate(speed = units::set_units(speed, "km/h")) %>%
  summary()
```
Hmm, 146 km/h is a bit fast for a boobie; they are known to reach ~90 km/h.

Let's plot the speed of each event:
```{r}
ggplot() +
  geom_event_path(
    data = boobies_sub,
    aes(color = speed)
  ) +
  facet_wrap(~bird_id)
```

Note that, as we did not save the object with the new speed units, the
original units are still in the object.

We can remove locations characterised by unlikely speeds using the McConnell's
algorithm (the same used by `trip::speedfilter()`). Note that we can give the
`max_speed` in any units we want, and they will be automatically converted to the
units in the object.

```{r}
boobies_sub <- boobies_sub %>%
  tt_clean_mcconnell(max_speed = units::as_units(90, "km/h"))
```

Note that `tt_clean_mcconnell()` is "destructive" and will remove the points
flagged by the algorithm. There is also a function `event_flag_mcconnell()`
which flags points without removing them; see the advanced vignette for more
details on how to use it (TODO there is no advance vignette yet!!!).

Let us now recompute the speeds (directly in km/h):

```{r}
boobies_sub <- boobies_sub %>%
  mutate(speed = event_speed(., units = "km/h"))
summary(boobies_sub)
```

We can see that we have removed step with an extreme speed. Note that the McConnell algorithm
does not simply remove steps above the threshold speed, but uses the root
mean square speed for previous/next and 2nd previous to next point, thus allowing
for individual steps that might have a speed just above the threshold if the subsequent
point is not too far away.


# Splitting trips

Boobies are central place foragers, and we now want to split the track into
trips. For this we need to define the colony location, and a buffer around it.
`tt_split_trips()` can take either a single colony point, or one nest location
per track. 

```{r}
boobies_sub <- boobies_sub %>%
  tt_split_trips(
    center_col = "colony_coord",
    buffer_outbound = as_units(3, "km"),
    buffer_inbound = as_units(10, "km")
  )
```

If we now inspect the object, we can see that there is a new column, `trip_id`,
which is also used as the new `track_id_column` in the `move2` object:
```{r}
boobies_sub
```

Let's look at the metadata:
```{r}
show_meta(boobies_sub)
```

We can see that only individual 69302 performed proper central place trips, and
specifically it made 3 trips. 69303 never moved, and 69304 migrated away from
the study area. We can now plot the
trips with:
```{r}
ggplot() +
  geom_sf(
    data = track_lines(boobies_sub),
    aes(color = trip_id)
  ) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

# Summarising tracks

We can get a summary of our tracks with:

```{r}
boobies_sub %>%
  track_summary_stats(center_col = "colony_coord")
```

Note that `track_summary_stats()` will return a tibble with one row per track,
rather than the usual vector of lenght equal to the number of tracks that is
produced by most `track_*` functions.

# Home Range estimates

We can now estimate the home range of individuals. First, we will process
the full dataset to split trips, and drop individuals that did not move
or migrated (such as the ones seen in the previous section):

```{r}
boobies_trips <- boobies_mt %>%
  tt_split_trips(
    center_col = "colony_coord",
    buffer_outbound = as_units(3, "km"),
    buffer_inbound = as_units(10, "km")
  )
```

Let's get a summary of the metadata of what is left:
```{r}
summary(show_meta(boobies_trips))
```

We'll take the 3 birds with the most trips:
```{r}
boobies_trips_sub <- boobies_trips %>%
  filter_by_meta(bird_id %in% c("69339", "69314", "69333"))
```
We can now plot the trips:
```{r}
ggplot() +
  geom_track_path(
    data = boobies_trips_sub,
    aes(color = trip_id)
  ) +
  coord_sf(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m") +
  facet_wrap(~bird_id) + theme(legend.position="none")
```


To estimate home ranges, we need to project the data to an equal area projection.
We can use the `sf::st_transform()` function to do this:
```{r}
boobies_trips_sub <- boobies_trips_sub %>%
  sf::st_transform(crs = "+proj=aeqd +lon_0=-16 +lat_0=-5 +units=m")
```

We can now estimate the home range using Kernel Density Estimation. We want to 
estimate the home range for each bird, so we will group by `bird_id`:

```{r}
boobies_kde <- boobies_trips_sub %>%
  group_by(bird_id) %>%
  tt_hr_kde(levels = c(0.5, 0.95))
```

`boobies_kde` is an `sf` object, with estimates of the area (in the units of
the projection, in this case m^2) and a geometry column:
```{r}
boobies_kde
```

We could easily change the units of the area with `set_units()`:
```{r}
boobies_kde %>%
  mutate(area = set_units(area, "km^2"))
```


We can plot the home ranges:
```{r}
ggplot(boobies_kde) +
  geom_sf(aes(fill = bird_id), alpha = 0.7)
```

We can simply add actual events with:
```{r}
ggplot() +
  geom_sf(
    data = boobies_kde,
    aes(fill = bird_id), alpha = 0.7
  ) +
  geom_event_point(
    data = boobies_trips_sub,
    size = 0.1
  ) +
  facet_wrap(~bird_id)
```

To get a minimum convex polygon covering 90% of the data, we can simply use:
```{r}
boobies_mcp <- boobies_trips_sub %>%
  group_by(bird_id) %>%
  tt_hr_mcp(levels = c(0.90))
```

We can plot the MCPs with:
```{r}
ggplot() +
  geom_sf(
    data = boobies_mcp,
    aes(fill = bird_id), alpha = 0.7
  ) +
  geom_event_point(
    data = boobies_trips_sub,
    size = 0.1
  ) +
  facet_wrap(~bird_id)
```
