---
title: "tidytrack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidytrack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidytrack)
```

We use a dataset on boobies from the `track2KBA` package. The dataset contains 
the tracking data of 43 boobies. We will use this dataset to demonstrate the 
functionalities of the `tidytrack` package.

```{r echo=FALSE, results='hide'}
# remove colony information from the dataset
boobies <- track2KBA::boobies %>% select(-c(lon_colony, lat_colony))
```

Let's inspect the table
```{r}
summary(boobies)
```

We start by combining the `data_gmt` and `time_gmt` columns into a single `POSIXct`,
using the `lubridate` package, and then removing the original columns. 
```{r}
library(dplyr)
library(lubridate)
boobies <- boobies %>%
  mutate(mt_time = ymd_hms(paste(date_gmt, time_gmt))) %>%
  select(-c(date_gmt, time_gmt))
head(boobies)
```

Next, we convert the data into an `sf` object.
```{r}
library(sf)
boobies_sf <- st_as_sf(boobies, coords = c("longitude", "latitude"),
                       crs = "+proj=longlat +datum=WGS84")
summary(boobies_sf)
```

We are now ready to cast the data into a `move2` object, which we will then
use as the base of our analysis:
```{r}
library(move2)
boobies_move <- mt_as_move2(boobies_sf, track_id_column = "track_id",
                            time_column = "mt_time")
boobies_move
```

Note that we have used the `track_id` column as the track identifier. This column
contains the ID of the bird that was tracked. The print method summarises quickly
the `move2` object for us. In a `move2` object, each row represents a timestamped
event, in this case GPS locations of the birds.
