---
title: "Testing UDs in adehabitatHR"
output: html_document
---

Here we are running through the functions for making and comparing utilization
distributions (UDs) in the R package `adehabitatHR` (Calenge 2006) so that we can
sort out the logic of the UD functions in `tidytracks` and compare the outputs.

## load packages

```{r load_packages}

library(adehabitatHR)

```

## load data

The adehabitatHR package has function for estimating home ranges from animal
tracking data.

The functions accept an object of class `SpatialPoints` (one animal) or
`SpatialPointsDataFrame` (multiple animals).

The example boar tracking data has environmental variables in `map` and 
tracking data in `relocs`. We're interested in the latter, which are stored 
as a `SpatialPointsDataFrame`.

```{r load_data}

data(puechabonsp)
class(puechabonsp)
names(puechabonsp)

# two spatial points data frames
summary(puechabonsp)

# the tracking locations are in relocs
head(as.data.frame(puechabonsp$relocs))

# map of elevation
image(puechabonsp$map, col=grey(c(1:10)/10))
# add relocs
plot(puechabonsp$relocs, add=TRUE,
     col=as.data.frame(puechabonsp$relocs)[,1])

```

## MCP

Compute 95% MCP for each animal. The output is stored as a 
`SpatialPolygonsDataFrame`.

```{r mcp}
# compute mcp using just the first column (coordinates) of relocs
mcp95 <- mcp(puechabonsp$relocs[,1], percent=95)

# spatial polygons data frame with area calculated
mcp95

# plot polygons with points on top
plot(mcp95)
plot(puechabonsp$relocs, add=TRUE)

```

## kernel UD

"An animal's use of space can be described by a bivariate probability density
function, the UD, which gives the probability density to relocate the animal at
any place according to the coordinates (x, y)"

The kernel method is used to estimate the UD using the tracking data.

"The basic principle of the method is the following: a bivariate kernel function
is placed over each relocation, and the values of these functions are averaged
together."

Bandwidth *h* choice:

* reference bandwidth: assumes a bivariate normal distribution, so often results 
in a too large h value, oversmoothing the UD.
* least-squares cross-validation (LSCV): minimizes the integrated square error 
between the estimated UD and the true UD. Can result in undersmoothing (too 
small h value) or failure to converge. 
* subjective visual choice (or track2KBA method for ecological relevance).

## `kernelUD` function

"The UD is estimated in each pixel of a grid superposed to the relocations."

This function takes either a `SpatialPoints` (one animal) or 
`SpatialPointsDataFrame` (multiple animals) object.

Parameters:

* `xy`: An object inheriting the class SpatialPoints containing the x and y 
relocations of the animal. If xy inherits the class SpatialPointsDataFrame, it 
should contain only one column (factor) corresponding to the identity of the 
animals for each relocation.
* `h`: a character string or a number - either the method to use or the h 
value to use. Default is `href`.
* `grid`: a number giving the size of the grid on which the UD should be 
estimated. Alternatively, this parameter may be an object inheriting the 
class SpatialPixels, that will be used for all animals. For the function 
kernelUD, it may in addition be a list of objects of class SpatialPixels, 
with named elements corresponding to each level of the factor id. Default is 60.
* `same4all`: logical. If TRUE, the same grid is used for all animals (and must 
be calculated within the function, so you can't use your own grid if this is TRUE). 
* `hlim`: some limits on LSCV bandwidth search (if h = "LSCV").
* `kern`: a character string. If "bivnorm", a bivariate normal kernel is used. 
If "epa", an Epanechnikov kernel is used.
* `extent`: value for how much to expand the grid around the range of 
relocations. Default is 1, i.e. grid roughly spans the data range.
* `boundary`: Default NULL. If, not NULL, an object inheriting the class 
SpatialLines defining a barrier that cannot be crossed by the animals.


?? If the UDs have been standardised, then the values should sum to 1. However,
they don't seem to!

## using href

```{r kernel_ud}

kud <- kernelUD(xy = puechabonsp$relocs[,1], # the coordinates field in the spdf
                h = "href") # using default params for everything else
kud
class(kud)

# plot for each animal
graphics::image(kud)

# what h values were used for each animal?
sapply(kud, function(x) x@h)

# do the values in each UD sum to 1? NO.
sapply(kud, function(x) sum(x@data))

```

The output is a list of class `estUDm` (estimated UD for multiple animals).

Let's look at the first element, the UD for Brock. It is class `estUD` and 
is a list containing objects in `@`, including:

* `@h`: the bandwidth used, and method
* `@data`: the values for each grid cell(?)
* `@grid`: a `GridTopology` object (from `sp` package) with 3 slots:
  * `cellcentre.offset`
  * `cellsize`
  * `cells.dim`

```{r kernel_ud_2}

kud_brock <- kud[["Brock"]]
class(kud_brock)

# h info
kud_brock@h

# first few values in data
head(kud_brock@data)
range(kud_brock@data)

# grid info
kud_brock@grid



#so to get grid for TT we want to reconstruct the bounding box
# also the resulution is cell size of 89.954 map units in each direction. So to get the grid we just need to reconstruct the bounding box and then split it into cells of 89.954 map units (how to work out if units the same?, does it matter? - actually probably not)

# find min and max coords of relocs before expanding by extent param
xmin <- min(as.data.frame(puechabonsp$relocs)[,2])
xmax <- max(as.data.frame(puechabonsp$relocs)[,2])
ymin <- min(as.data.frame(puechabonsp$relocs)[,3])
ymax <- max(as.data.frame(puechabonsp$relocs)[,3])
```

## using LCSV

```{r kernel_ud_lcsv}
# run the kernel UD
kudl <- kernelUD(puechabonsp$relocs[,1], h="LSCV")

# look at the resulting UDs
image(kudl)

# look at the results of the LCSV minimisation to see if they converged
plotLSCV(kudl)

# what h values were used for each animal?
sapply(kudl, function(x) x@h)
```

## controlling the grid

The UD is estimated at the center of each pixel of a grid. Although the size
and resolution of the grid does not have a large effect on the estimates (see
Silverman, 1986), it is sometimes useful to be able to control the parameters
defining this grid. This is done thanks to the parameter `grid` of the function
`kernelUD`.

### numeric value

When a numeric value is passed to the parameter grid, the function `kernelUD`
automatically computes the grid for the estimation. This grid is computed in
the following way:

* the minimum and maximum coordinates (`Xmin`, `Xmax`, `Ymin`, `Ymax`) of the
relocations are computed.
* the range of `X` and `Y` values is also computed (i.e., the difference between
the maximum and minimum value). Let `Rx` and `RY` be these ranges.
* The coverage of the grid is defined thanks to the parameter extent of
the function kernelUD. Let `e` be this parameter. The minimum and maximum 
`X` coordinates of the pixels of the grid are equal to `Xmin − e × Rx`
and `Xmax + e × Rx` respectively. Similarly, the minimum and maximum
`Y` coordinates of the pixels of the grid are equal to `Ymin − e × RY` and
`Ymax + e × RY` respectively.
* Finally, these extents are split into `g` intervals, where `g` is the value 
of the parameter `grid.`



Consequently, the parameter `grid` controls the resolution of the grid and the
parameter extent controls its extent. 

Using Brock to illustrate:

```{r kernel_ud_grid_numeric}

## The relocations of "Brock"
locs <- puechabonsp$relocs
firs <- locs[as.data.frame(locs)[,1]=="Brock",]
## Graphical parameters
par(mar=c(0,0,2,0))
par(mfrow=c(2,2))
## Estimation of the UD with grid=20 and extent=0.2
image(kernelUD(firs, grid=20, extent=0.2))
title(main="grid=20, extent=0.2")
## Estimation of the UD with grid=80 and extent=0.2
image(kernelUD(firs, grid=80, extent=0.2))
title(main="grid=80, extent=0.2")
## Estimation of the UD with grid=20 and extent=3
image(kernelUD(firs, grid=20, extent=3))
title(main="grid=20, extent=3")
## Estimation of the UD with grid=80 and extent=3
image(kernelUD(firs, grid=80, extent=3))
title(main="grid=80, extent=3")

## extract grid info for one of the UDs
kud_brock_20_02 <- kernelUD(firs, grid=20, extent=0.2)


kud_brock_20_02@grid # gives us the grid topology, including the cell size and dimensions of the grid

library(sf)
sf::st_crs(puechabonsp$relocs)


```

### same for all

By default, the grid is computed separately for each animal. However, it is
possible to use the same grid for all animals by setting the parameter
`same4all` to TRUE.

```{r kernel_ud_grid_same4all}

kus <- kernelUD(puechabonsp$relocs[,1], same4all=TRUE)
image(kus)
# do the values in each UD sum to 1? NO.
sapply(kus, function(x) sum(x@data))

```

### SpatialPixelsDataFrame

If you have a raster map already available (e.g. of environmental data), you 
can use that as the grid.

In the example data. there are environmental layers under `puechabonsp$map`.

```{r kernel_ud_grid_spdf}
kudm <- kernelUD(puechabonsp$relocs[,1], grid=puechabonsp$map)
image(kudm)
```

## isopleth as vector

NB. isopleth us referred to in the adehabitatHR documentation as "home range".

The `getverticeshr` function extracts the isopleths from a UD object. It works 
on either a `estUD` (one animal) or `estUDm` (multiple animals) object.

NB. this function runs `getvolumeUD` internally to compute the cumulative UD, 
see: https://github.com/cran/adehabitatHR/blob/master/R/getverticeshr.r

```{r isopleths_vectors}

iso_90 <- getverticeshr(x = kudl,
                        percent = 90)
class(iso_90)
plot(iso_90, col=1:4)

```

## cumulative UD

`kernelUD` makes the UD (continuous surface probability distribution, in which 
all values sum to 1). 

`getvolumeUD`  ranks cells in the UD by cumulative probability (converts density
values to cumulative volume values such that cell values range from 0 to 1, i.e.
"what proportion of the total UD is contained in areas with density >= this
cell?")

```{r cumulative_ud}

vud <-  getvolumeUD(kudl)
vud
class(vud[[1]])

sum(vud[[1]]@data)

```

Plot both kernel UD and volume UD for first animal (Brock):

```{r plot_kud_vud}

## Set up graphical parameters
par(mfrow=c(2,1))
par(mar=c(0,0,2,0))
## The output of kernelUD for the first animal
image(kudl[[1]])
title("Output of kernelUD")
## Convert into a suitable data structure for
## the use of contour
xyz <- as.image.SpatialGridDataFrame(kudl[[1]])
contour(xyz, add=TRUE)
## and similarly for the output of getvolumeUD
par(mar=c(0,0,2,0))
image(vud[[1]])
title("Output of getvolumeUD")
xyzv <- as.image.SpatialGridDataFrame(vud[[1]])
contour(xyzv, add=TRUE)

```

Whereas the output of `kernelUD` is the raw UD, the output of `getvolumeUD` can
be used to compute the home range (the labels of the contour lines correspond to
the home ranges computed with various probability levels). For example, to get
the rasterized 95% home range of the first animal:

```{r rasterized_hr}
## store the volume under the UD (as computed by getvolumeUD)
## of the first animal in fud
fud <- vud[[1]]
## store the value of the volume under UD in a vector hr95
hr95 <- as.data.frame(fud)[,1]
## if hr95 is <= 95 then the pixel belongs to the home range
## (takes the value 1, 0 otherwise)
hr95 <- as.numeric(hr95 <= 95)
## Converts into a data frame
hr95 <- data.frame(hr95)
## Converts to a SpatialPixelsDataFrame
coordinates(hr95) <- coordinates(vud[[1]])
gridded(hr95) <- TRUE
## display the results
image(hr95)

```


## isopleth area

"home range area" is computed automatically by `getverticeshr`. As for the MCP,
the units of these areas are controlled by the parameters `unin` and `unout` of
the function `getverticeshr`

```{r iso_area}
as.data.frame(iso_90)
```

To compute isopleth area for several levels, you can use `kernel.area` function:

```{r kernel_area}
iso_areas <- kernel.area(kudl, percent=seq(50, 95, by=5))
class(iso_areas)
iso_areas
plot(iso_areas)
```

## extensions

These are discussed in the adehabitatHR vignette, but I don't think we need to 
implement them in `tidytracks`, at least not yet.

### physical boundaries

here in the vignette there is some stuff about how to impose a physical 
boundary on the UD, e.g. if there is a pond or river and the animal does not 
go into water. I don't think we need to implement this in tidytracks right now.

### bridge kernel methods

Instead of placing a kernel function over each relocation, this method places a
kernel function over each movement step (straight line between successive
locations). This takes into account the path travelled by the animal AND the
time that it took, and so gets around some of the issues with autocorrelation in
tracking data.

**Brownian bridge kernel method**

This is implemented in the `kernelbb` function in adehabitatHR, and requires two 
parameters for the kernel functions. It also requires the tracking data to be in 
the form of a `ltraj` object from the `adehabitatLT` package.

There is also the **biased random bridge method**, in the function `BRB`,
followed in the adehabitatHR vignette by a discussion of "decomposing the use of
space into intensity and recursion", i.e. distinguishing between the places that
the animal often uses and the places that are used only once, but for a long
time.

### adding time

The 'product kernel algorithm' includes time as a third dimension.

## convex hulls and LoCoH

Also discussed in the vignette and implemented in adehabitatHR, but not the 
focus of this document.
