---
title: "Animations with tidytracks"
output: html_document
date: "2026-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE}
library(tidytracks)
library(ggplot2)
library(sf)
library(terra)
library(tidyterra)
library(rnaturalearth)
library(gganimate)
```


`tidytracks` makes creating animated maps of your animal movement data easy.

First, load your data. Here we are using 10 wandering albatross tracks from 
Bird Island, South Georgia, each lasting approximately 1 year.

```{r load_data}
# load data from CSV file into tidytracks
df <- tt_read_data(events = "wanderer_example_tracks.csv",
                   col_track_id = "bird_id",
                   col_coords = c("lon", "lat"),
                   col_date_time = "date_time")

# inspect the data
df
```

You need regular timestamps to animate, so regularise your timestamps.

```{r interpolate}
df <- move2::mt_interpolate(x = df,
                            time = "12 hours",
                            omit = TRUE)

df
```


Read in any other layers you want to include in your map, e.g.

* raster file such as bathymetry to use as a basemap
* polygons e.g. countries, national park boundaries
* points e.g. colony or tagging location

```{r load_extra_layers}
# # load bathymetry raster using terra package
# bathy <- terra::rast("bathy_capped.nc")
# # transform it to WGS84 to match df
# bathy <- terra::project(bathy, "EPSG:4326")
# bathy

# make colony dataframe, make it spatial and add CRS using sf package
colony <- data.frame(colony = "Bird Island", lon = -38.03, lat = -54) %>%
  sf::st_as_sf(coords = c("lon", "lat")) %>%
  sf::st_set_crs(4326)

# read countries from rnaturalearth at medium res
land <- rnaturalearth::ne_countries(scale = "medium",
                                    returnclass = "sf") %>%
  dplyr::select(name, continent)
```

Sort out your projections: check that the CRS of all your layers match, and if 
needed, transform every object to your desired projection for plotting.

```{r sort_projections}
# check CRS of df matches the other layers for the plot
st_crs(df) == st_crs(colony)
# st_crs(df) == st_crs(bathy)
st_crs(df) == st_crs(land)

# define south polar LAEA projection for plotting
crs_laea_south <- "+proj=laea +lat_0=-90 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"

# transform all my layers into the new projection
df_p <- sf::st_transform(df, crs_laea_south)
colony_p <- sf::st_transform(colony, crs_laea_south)
land_p <- sf::st_transform(land, crs_laea_south)
# bathy_p <- terra::project(bathy, crs_laea_south)
```

Next, assemble your map using `ggplot2`. Add `sf` layers using `geom_sf`, 
`terra` layers using `tidyterra`'s functions `geom_spatraster` (for rasters) or 
`geom_spatvector` (for vectors), and add your `tidytracks` layer using 
`geom_event_path` for paths or `geom_event_point` for points.

```{r assemble_map}

map <- ggplot() +
  # add land polygons
  geom_sf(data = land_p, fill = "grey50", col = NA) +
  # add tidytracks layer
  # geom_event_point(data = df_p, aes(col = bird_id), size = 1.5) +
  geom_event_path(data = df_p, aes(col = bird_id), size = 1.5) +
  # add colony layer
  geom_sf(data = colony_p, pch = 24, fill = "yellow") +
  # update extent of map using extent of tracking data
  coord_sf(xlim = c(min(st_coordinates(df_p)[, 1]),
                    max(st_coordinates(df_p)[, 1])),
           ylim = c(min(st_coordinates(df_p)[, 2]),
                    max(st_coordinates(df_p)[, 2]))
  )

map
```

Animate the map. This produces a list of 2 elements. The first is the 

```{r animate_map}
list_anim <- tt_animate_map(p = map, 
                            x = df_p,
                            wake_length = 0.1,
                            label_format = "%B %Y")

class(list_anim)
names(list_anim)
class(list_anim$p_anim)
class(list_anim$n_timesteps)
list_anim$n_timesteps
```

Render the animation. By default, gganimate uses 100 frames. If you want one 
frame of the animation per timestep in your original dataframe, use 
`nframes = list_anim$n_timesteps`.

TODO I think nframes needs to be at least as many as, if not more than, the 
number of timesteps in your data, for animating PATHS to work.

```{r animate_map_2}

animate(plot = list_anim$p_anim,
        nframes = list_anim$n_timesteps,
        duration = 20)

```

