---
title: "Animations with tidytracks"
output: 
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE}
library(tidytracks)
library(ggplot2)
library(sf)
library(gganimate)
```


`tidytracks` makes creating animated maps of your animal movement data easy.

<br>

***

## load tracking data

First, load your data. Here we are using 10 wandering albatross tracks from 
Bird Island, South Georgia, each lasting approximately 1 year.

```{r load_data, collapse=TRUE}
# load data from CSV file into tidytracks
df <- tt_read_data(events = "wanderer_example_tracks.csv",
                   col_track_id = "bird_id",
                   col_coords = c("lon", "lat"),
                   col_date_time = "date_time")

# inspect the data
df
```

### regularise data

Animation works best with regular timesteps and, if you have multiple 
individuals, matching timestamps across individuals that were tracked 
concurrently. Therefore we regularise our timestamps if they are not already 
regular.

```{r interpolate, collapse=TRUE}
df <- move2::mt_interpolate(x = df,
                            time = "12 hours",
                            omit = TRUE)

df
```

<br>

***

## example 1: map with polygons

### load layers

Read in any other layers you want to include in your map, e.g.

* raster file such as bathymetry to use as a basemap
* polygons e.g. countries, national park boundaries
* points e.g. colony or tagging location

```{r load_extra_layers}

# make colony dataframe, make it spatial and add CRS using sf package
colony <- data.frame(colony = "Bird Island", lon = -38.03, lat = -54) %>%
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326)

# read countries from rnaturalearth at medium res
library(rnaturalearth)
land <- ne_countries(scale = "medium", returnclass = "sf")
```

### projections

Sort out your projections: check that the CRS of all your layers match, and if 
needed, transform every object to your desired projection for plotting.

```{r sort_projections, collapse=TRUE}
# check CRS of df matches the other layers for the plot
st_crs(df) == st_crs(colony)
st_crs(df) == st_crs(land)

# define south polar LAEA projection for plotting
my_crs <- "+proj=laea +lat_0=-90 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"

# transform all my layers into the new projection
df_p <- sf::st_transform(df, my_crs)
colony_p <- sf::st_transform(colony, my_crs)
land_p <- sf::st_transform(land, my_crs)
```

### make map

Next, assemble your map using `ggplot2`. Add `sf` layers using `geom_sf`, 
`terra` layers using `tidyterra`'s functions `geom_spatraster` (for rasters) or 
`geom_spatvector` (for vectors), and add your `tidytracks` layer using 
`geom_event_path` for paths or `geom_event_point` for points.

```{r assemble_map}
map <- ggplot() +
  # geom_event_point(data = df_p, aes(col = bird_id), size = 1.5) +
  geom_event_path(data = df_p, aes(col = bird_id), size = 1.5) + # tracks layer
  geom_sf(data = land_p, fill = "grey50", col = NA) + # land polygons
  geom_sf(data = colony_p, pch = 24, fill = "yellow") # add colony layer

map
```

Use `ggplot2` functions `coord_cartesian` or `coord_sf` to change the limits 
of the plot (alternatively, you can crop all your layers to the required 
area before making the plot).

```{r, collapse=TRUE}
# you can use st_coordinates to get min/max coordinates of your tracking data
# NB. this will be in the projected units, not necessarily lon/lat
min(st_coordinates(df_p)[, 1]) # example: min x coordinate

# update extent of map using extent of tracking data
map <- map +
  coord_sf(xlim = c(min(st_coordinates(df_p)[, 1]),
                    max(st_coordinates(df_p)[, 1])),
           ylim = c(min(st_coordinates(df_p)[, 2]),
                    max(st_coordinates(df_p)[, 2])))
map
```

### animate map

Animate the map using `animate_map()`. This produces a `list` of 2 elements.

* `p_anim` is the `gganim` object (a `ggplot2` object with extra animation 
logic). We will use the `animate` function from the `gganimate` package to 
render the animation from this object.
* `n_timesteps` is the number of unique timestamps in the tracking data used to
make the plot. To ensure a smooth animation, we need to render it with **at
least as many frames as there are timesteps** in the tracking data.

```{r animate_map, collapse=TRUE}
list_anim <- animate_map(p = map, 
                         x = df_p,
                         wake_length = 0.1,
                         label_format = "%B %Y")

class(list_anim) # the output of animate_map is a list
names(list_anim) # what are the elements in the list?
class(list_anim$p_anim) # this is the gganim object
list_anim$n_timesteps # this is the number of unique timesteps
```

### render animation

Now, we render the animation using `gganimate::animate()`. By default,
`gganimate` uses 100 frames. If you want one frame per timestep, use 
`nframes = list_anim$n_timesteps`.

You can **specify the length of your animation** using two out of these three 
parameters:

* `nframes`: number of frames to render (default 100)
* `fps`: framerate in frames/second (default 10)
* `duration`: duration of the animation in seconds (no default)

Here we use the suggested number of frames, and a duration of 20 seconds. 
Depending on the number of frames, this may take a few minutes or longer.

```{r animate_map_2, message=FALSE, warning=FALSE}
gganimate::animate(
  plot = list_anim$p_anim,
  nframes = list_anim$n_timesteps,
  duration = 20
)
```

To save the animation to file, use `gganimate`'s `anim_save()` function which 
by default saves the most recent animation, similarly to `ggsave()`.

```{r eval=FALSE}
anim_save("example_animation.gif")
```

### points vs paths

This animation included the tracking data as paths using `geom_event_path()`. 
Let's illustrate the difference between paths and points by re-creating it 
using `geom_event_point()`.

```{r full_anim_points, message=FALSE, warning=FALSE}
# first, make the map and print it to check
map <- ggplot() +
  geom_event_point(data = df_p, aes(col = bird_id), size = 1.5) + # tracks layer
  geom_sf(data = land_p, fill = "grey50", col = NA) + # land polygons
  geom_sf(data = colony_p, pch = 24, fill = "yellow") + # add colony layer
  coord_sf(xlim = c(min(st_coordinates(df_p)[, 1]), # define extents
                    max(st_coordinates(df_p)[, 1])),
           ylim = c(min(st_coordinates(df_p)[, 2]),
                    max(st_coordinates(df_p)[, 2])))
# print to check
map
# add animation
list_anim <- animate_map(p = map, 
                         x = df_p,
                         wake_length = 0.1,
                         label_format = "%B %Y")
# render the animation
gganimate::animate(
  plot = list_anim$p_anim,
  nframes = list_anim$n_timesteps,
  duration = 20
)
```

Points animations work best when the resolution of the tracking data is high, 
so points are closer together.

```{r include=FALSE}
# clean up from this section
rm(map, list_anim)
```

<br>

***

## example 2: map with raster

In this example we're going to add a basemap as a `spatRaster` object using 
the package `terra`. We'll download a raster using the `basemaps` package, 
but you can also load in your own raster (for example, from a `geotiff` or 
`netCDF` file) using `terra::rast()` and add it to the map using 
`tidyterra::geom_spatraster`.

```{r, message=FALSE, warning=FALSE}
# load packages for acquiring, loading, and plotting rasters
library(basemaps)
library(terra)
library(tidyterra)
```

### get basemap

To download using `basemaps`, you need to ask for a defined spatial extent 
(in longitude/latitude) - usually we want the bounding box of our data so we 
can define the extent programmatically, but sometimes you may want to define 
it manually, or add a buffer around your data.

```{r, collapse=TRUE}
# make a rectangle as the bounding box of all the data (in lon/lat)
rect <- st_as_sfc(st_bbox(df))
rect

# I actually want to extend my rectangle to the whole Southern Ocean
# and up to the northern extent of my data, plus a 15 degree buffer
rect_coords <- c(xmin = -180, # define the corners
                 xmax = 180,
                 ymin = -90,
                 ymax = max(st_coordinates(df)[, 2]) + 15) # add buffer
# convert to an sfc object and set the CRS
rect <- st_as_sfc(st_bbox(rect_coords)) %>%
  st_set_crs(4326)
rect

```

Now we have the extent `rect`, let's download the basemap, check its CRS, and 
project it to the projection we're using for our map.

```{r warning=FALSE, collapse=TRUE}
# download basemap as a terra spatRaster
bmap <- basemap_terra(rect, map_service = "esri", map_type = "world_ocean_base")
# check the CRS it comes in
terra::crs(bmap, describe = TRUE)
# re-project basemap to our custom projection
bmap_p <- terra::project(bmap, my_crs)

# inspect the raster: it is in RGB (red, green, blue) format
bmap_p
```

### make map

Now assemble our map. We can add the raster layer using the `tidyterra`
functions `geom_spatraster` (for single layer e.g. bathymetry raster) or
`geom_spatraster_rgb` (for RGB rasters with a layer for each of red, gree,
blue).

```{r}
# use different colour scale on top of RGB raster
track_colours <- c(
  "#FFD700", # Gold
  "#FF6B35", # Orange
  "#FF3399", # Hot pink
  "#39FF14", # Neon green
  "#FF4444", # Red
  "#00FFCC", # Aqua
  "#CC44FF", # Purple
  "#FFFEE4", # White ish
  "#FF7070", # Coral
  "#CFFF04" # Neon yellow
)

# assemble map
map <- ggplot() +
  geom_spatraster_rgb(data = bmap_p) + # basemap layer first
  geom_event_path(data = df_p, aes(col = bird_id), size = 1.5) + # tracks layer
  scale_colour_manual(values = track_colours) + # change track colours
  geom_sf(data = colony_p, pch = 24, fill = "yellow") # add colony layer
map

# remove axes and gridlines - looks neater
map <- map +
  ggplot2::theme_void()
map
```

### animate

Now add the animation logic to the map and render the animation with the 
suggested number of frames.

This time we'll add the animation by piping the `ggplot` to `animate_map` - 
both ways of using the function work.

```{r, message=FALSE, warning=FALSE}
# add animation to the map, this time with pipe 
list_anim <- map %>%
  animate_map(.,
              x = df_p,
              wake_length = 0.1,
              label_format = "%B %Y")

# render the animation
gganimate::animate(
  plot = list_anim$p_anim,
  nframes = list_anim$n_timesteps,
  duration = 20
)
```

<br>

***

