---
title: "Animating your maps"
output: 
  rmarkdown::html_vignette
#  html_document
vignette: >
  %\VignetteIndexEntry{Animating your maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`tidytracks` also includes a function for animating maps of animal movement 
data over time.

```{r load_packages, message=FALSE, warning=FALSE}
library(tidytracks)
library(ggplot2)
library(sf)
library(gganimate)
```

<br>

***

## load tracking data

First, load your data. Here we are using 10 wandering albatross tracks from 
Bird Island, South Georgia, each lasting approximately 1 year.

```{r load_data, collapse=TRUE}
# load data from CSV file into tidytracks
df <- tt_read_data(events = "wanderer_example_tracks.csv",
                   col_track_id = "bird_id",
                   col_coords = c("lon", "lat"),
                   col_date_time = "date_time")

# inspect the data
df
```

### regularise data

Animation works best with regular time steps and, if you have multiple 
individuals, matching timestamps across individuals that were tracked 
concurrently. Therefore we regularise our timestamps if we haven't already:

```{r interpolate, collapse=TRUE}
df <- tt_regular_time(x = df,
                      interval = as_units(12, "hour"),
                      snap_times = TRUE)
df
```

<br>

***

## examples 1 & 2: map with polygons

NB. We use a subset (first 30 days) of the dataset for these maps to keep the
file size of the animations low.

### load layers

Read in any other layers you want to include in your map, e.g.

* raster file such as bathymetry to use as a basemap
* polygons e.g. countries, national park boundaries
* points e.g. colony or tagging location

```{r load_extra_layers}
# make colony dataframe, make it spatial and add CRS using sf package
colony <- data.frame(colony = "Bird Island", lon = -38.03, lat = -54) %>%
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326)

# read countries from rnaturalearth at medium res
library(rnaturalearth)
land <- ne_countries(scale = "medium", returnclass = "sf")
```

### subset data to test

We're going to test out our maps and animations on the first 30 days of the 
data.

```{r}
# subset the tracking data to the first 30 days, to test our animation
df_sub <- df %>%
  dplyr::filter(date_time <= (min(df$date_time) + lubridate::days(30)))
```

### make map

Next, assemble your map using `ggplot2`. Add `sf` layers using `geom_sf`, 
`terra` layers using `tidyterra`'s functions `geom_spatraster` (for rasters) or 
`geom_spatvector` (for vectors), and add your `tidytracks` layer using 
`geom_event_path` for paths or `geom_event_point` for points.

```{r assemble_map}
# make the map
map <- ggplot() +
  geom_event_path(data = df_sub, aes(col = bird_id),  # tracks layer
                  size = 1.5, lineend="round") +  # rounded line ends look smoother
  geom_sf(data = land, fill = "grey50", col = NA) + # land polygons
  geom_sf(data = colony, pch = 24, fill = "yellow") # add colony layer
# print the map
map
```

### projection and limits

Use `ggplot2` functions `coord_cartesian` or `coord_sf` to change the limits 
of the plot (alternatively, you can crop all your layers to the required 
area before making the plot). `coord_sf` can also change the projection.

```{r map_add_proj_lims, collapse=TRUE, warning=FALSE}
# define south polar LAEA projection for plotting
my_crs <- "+proj=laea +lat_0=-90 +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs"

# you can use st_coordinates to get min/max coordinates of your tracking data
range(st_coordinates(df)[ ,"X"])
range(st_coordinates(df)[ ,"Y"])

# update extent of map using extent of tracking data
map <- map +
  coord_sf(crs = my_crs, # change projection to LAEA South Polar
           # NB. lims must be in the units of the target CRS, so we need to
           # transform the data to that crs using st_transform
           xlim = range(st_coordinates(st_transform(df_sub, my_crs))[ ,"X"]),
           ylim = range(st_coordinates(st_transform(df_sub, my_crs))[ ,"Y"]))
map
```


### animate map

Animate the map using `animate_map()`. This produces a `gganim` object (a
`ggplot2` object with extra animation logic). We will use the `animate` function
from the `gganimate` package to render the animation from this object.

Our `gganim` object has an attribute called `n_timesteps`. This is the number of
unique timestamps in the tracking data used to make the plot. To ensure a smooth
animation, we need to render it with **at least as many frames as there are
time steps** in the tracking data.

```{r animate_map, collapse=TRUE, warning=FALSE}
map_anim <- animate_map(p = map,           # our ggplot map
                        wake_length = 0.3, # proportion of total animation time
                        label_format = "%B %d" # month and date
) 

class(map_anim)               # the gganim object
attr(map_anim, "n_timesteps") # number of time steps in animation
```

### render animation

Now, we render the animation using `gganimate::animate()`. By default,
`gganimate` uses 100 frames. We will change this using the `nframes` parameter.

You can **specify the length of your animation** using two out of these three 
parameters:

* `nframes`: number of frames to render (default 100)
* `fps`: framerate in frames/second (default 10)
* `duration`: duration of the animation in seconds (no default)

Here we use the suggested number of frames, and a duration of 20 seconds. 
Depending on the number of frames, rendering take a few minutes or longer.

See the [gganimate documentation](https://gganimate.com/reference/animate.html)
for more information on the different renderer options for making your
animation. You can also save each frame as an image file and use your own
software to bind them into a video.

```{r animate_map_2, message=FALSE, warning=FALSE, fig.width=5.4, fig.height=4}
gganimate::animate(
  plot = map_anim,         # our ggplot with added animation
  nframes = attr(map_anim, "n_timesteps"), # number of frames = time steps in data
  duration = 5,            # video duration in seconds
  renderer = av_renderer(), # default is gifski_renderer()
  # optionally, specify width and height of output animation
  width = 540, height = 400,
  units = "px", res = 96 # w/h in pixels, res in dpi
)
```

To save the animation to file, use `gganimate`'s `anim_save()` function which 
by default saves the most recent animation, similarly to `ggsave()`.

```{r eval=FALSE}
anim_save("example_animation.webm") # add appropriate file extension
```

```{r include=FALSE}
rm(map, map_anim) # clean up
```


### points vs paths

This animation included the tracking data as paths using `geom_event_path()`. 
Let's illustrate the difference between paths and points by re-creating it 
using `geom_event_point()`.

```{r anim_points, message=FALSE, warning=FALSE}
# first, make the map and print it to check
map <- ggplot() +
  geom_event_point(data = df_sub, aes(col = bird_id), size = 3) + # tracks layer
  geom_sf(data = land, fill = "grey50", col = NA) + # land polygons
  geom_sf(data = colony, pch = 24, fill = "yellow") + # add colony layer
  coord_sf(crs = my_crs, # change projection and define limits
           xlim = range(st_coordinates(st_transform(df_sub, my_crs))[ ,"X"]),
           ylim = range(st_coordinates(st_transform(df_sub, my_crs))[ ,"Y"]))
# print to check
map
# add animation
map_anim2 <- animate_map(p = map, 
                         wake_length = 0.3,
                         label_format = "%B %d")
```

```{r anim_points_render, message=FALSE, warning=FALSE, fig.width=5.4, fig.height=4}
# render the animation
gganimate::animate(
  plot = map_anim2,
  nframes = attr(map_anim2, "n_timesteps"),
  duration = 5,
  renderer = av_renderer(),
  # optionally, specify width and height of output animation
  width = 540, height = 400,
  units = "px", res = 96
)
```

Points animations work best when the resolution of the tracking data is high, 
so points are closer together.

```{r include=FALSE}
# clean up from this section
rm(map, map_anim2)
```

<br>

***

## example 3: map with raster

In this example we're going to add a basemap as a `spatRaster` object using 
the package `terra`. We'll download a raster using the `basemaps` package, 
but you can also load in your own raster (for example, from a `geotiff` or 
`netCDF` file) using `terra::rast()` and add it to the map using 
`tidyterra::geom_spatraster`.

```{r, message=FALSE, warning=FALSE}
# load packages for acquiring, loading, and plotting rasters
library(basemaps)
library(terra)
library(tidyterra)
```

### get basemap

To download using `basemaps`, you need to ask for a defined spatial extent 
(in longitude/latitude) - usually we want the bounding box of our data so we 
can define the extent programmatically, but sometimes you may want to define 
it manually, or add a buffer around your data.

```{r make_bbox, collapse=TRUE}
# option 1: make a rectangle as the bounding box of all the data (in lon/lat)
rect <- st_as_sfc(st_bbox(df))

# option 2: specify the rectangle manually. I want the whole Southern Ocean
# and up to the northern extent of my data (plus a 15 degree buffer)
rect_coords <- c(xmin = -180, # define the corners
                 xmax = 180,
                 ymin = -90,
                 ymax = max(st_coordinates(df)[, "Y"]) + 15) # add buffer
# convert to an sfc object and set the CRS
rect <- st_as_sfc(st_bbox(rect_coords)) %>%
  st_set_crs(4326)
rect
```

Now we have the extent `rect`, let's download the basemap.

```{r download_bmap, warning=FALSE, collapse=TRUE}
# download basemap as a terra spatRaster
bmap <- basemap_terra(rect, map_service = "esri", map_type = "world_ocean_base")
```

### make map

Now assemble our map. We can add the raster layer using the `tidyterra`
functions `geom_spatraster` (for single layer e.g. bathymetry raster) or
`geom_spatraster_rgb` (for RGB rasters with a layer for each of red, green,
blue).

```{r make_raster_map}
# define new track colours to show up better on a blue background
track_colours <- c(
  "#FFD700", # Gold
  "#FF6B35", # Orange
  "#FF3399", # Hot pink
  "#39FF14", # Neon green
  "#FF4444", # Red
  "#00FFCC", # Aqua
  "#CC44FF", # Purple
  "#FFFEE4", # White ish
  "#FF7070", # Coral
  "#CFFF04" # Neon yellow
)

# assemble map
map <- ggplot() +
  geom_spatraster_rgb(data = bmap) +                       # basemap
  geom_event_path(data = df, aes(col = bird_id), size = 2, # tracks
                  lineend = "round") + 
  scale_colour_manual(values = track_colours) +            # track colours
  geom_sf(data = colony, pch = 24, fill = "yellow") +      # colony
  coord_sf(crs = my_crs)                                   # projection

# remove axes and gridlines so it looks neater
map <- map +
  theme_void() +
  theme(
    plot.background = element_rect(
      fill = "white",
      colour = "white"))
map
```

### animate

Now add the animation logic to the map and render the animation with the 
suggested number of frames.

This time we'll add the animation by piping the `ggplot` to `animate_map` - 
both ways of using the function work.

```{r animate_raster_map, message=FALSE, warning=FALSE, fig.width=5.7, fig.height=5}
# add animation to the map, this time with pipe 
map_anim <- map %>%
  animate_map(wake_length = 0.1,
              label_format = "%B %Y")

# render the animation
gganimate::animate(
  plot = map_anim,
  nframes = attr(map_anim, "n_timesteps"),
  duration = 20,
  renderer = av_renderer(),
  # optionally, specify width and height of output animation
  width = 568, height = 500,
  units = "px", res = 96
)
```

<br>

***

