---
title: "Checking_tt_regular_time"
author: "HP"
date: "2026-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Trying tt_reg_time on some example data then test same data with the old function in dev branch
### Load a couple of packages and check the documentation of the function

```{r}
library(dplyr)
library(ggplot2)

?tt_regular_time
```


### Load some package example data
It's already a tidytracks object, can check this by inspecting metadata too

```{r}
load("data/shags_tt.rda")
shags_tt

head(show_meta(shags_tt))

```

Plot the data to see what it looks like

```{r}

ggplot() +
  geom_event_point(data = shags_tt, aes(color = bird_id))
```

 Needs filtering, remove low latitude point and plot again and project on the fly
 
``` {r}
shags_tt <- shags_tt %>%
  dplyr::filter(sf::st_coordinates(geometry)[,2] < -5)

ggplot() +
  geom_event_point(data = shags_tt, aes(color = bird_id)) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs')
```

Get idea of intervals for each shag track before resampling

```{r}
shags_tt %>%
  group_by(bird_id) %>%
  sf::st_drop_geometry() %>%
  summarise(median_dt = median(diff(date_time)),
            mean_dt = mean(diff(date_time)),
            min_dt = min(diff(date_time)),
            max_dt = max(diff(date_time)))%>%
  ungroup()
```

Just pick 2 bird IDs to resample, with different frequencies and track types. 
 No. 19 has longer gaps (approx 10 mins), 45 has shorter gaps (approx 2 mins)
 
```{r}
shags_tt_sub <- shags_tt %>%
  dplyr::filter(bird_id %in% c("KB_22_42657_19", "KB_23_42778_45"))

```

Now plot original tracks to compare later

```{r}
plot_1 <- ggplot() +
  geom_event_point(data = shags_tt_sub, aes(color = bird_id)) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs')+
  ggtitle("Original")

```
### Now resample the data using tt_regular_time

```{r}
# check that timestamps are posixt - yep

class(shags_tt_sub$date_time)

# mean intervals quite different, choose something between the both - 5 min?

resampled_shags <- tt_regular_time(shags_tt_sub,
                                   interval = units::set_units(5, "min"),
                                   max_time_lag = NULL)
```

Plot the resampled data and compare to the original

```{r}
plot_2 <- ggplot() +
  geom_event_point(data = resampled_shags, aes(color = bird_id)) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs')+
  ggtitle("Resampled")#

patchwork::wrap_plots(plot_1, plot_2, guides = "collect") 

```

# look at time differences in resampled data

```{r}
resampled_shags %>%
  group_by(bird_id) %>%
  sf::st_drop_geometry() %>%
  summarise(median_dt = median(diff(date_time)),
            mean_dt = mean(diff(date_time)),
            min_dt = min(diff(date_time)),
            max_dt = max(diff(date_time)))%>%
  ungroup()
```

Good.

## now trying the simple old version from dev branch

Wasn't confident of getting right version switching between branches so just 
using function directly here, renamed as tt_regular_time_dev

```{r}
tt_regular_time_dev <- function(x, every) {
  # replace minutes with mins in every if present
  every <- gsub("minutes", "mins", every)
  # same for seconds and secs
  every <- gsub("seconds", "secs", every)
  # now interpolate using move2
  return(move2::mt_interpolate(x, time = every, omit = TRUE))
}
```

Run on shag data

```{r}
resampled_shags_dev <- tt_regular_time_dev(shags_tt_sub,
                                   every = "5 mins")

head(resampled_shags_dev)
```

Inspect the time diffs and data. Main difference is that 
the old version snaps to the nearest 5 mins while the new version
starts with the first point and samples every 5 mins from there. The new
version results in a marginally higher number of observations, apparently regardless 
of interval set (have tested on a few different)

## Plot all and compare

```{r}

plot_3 <- ggplot() +
  geom_event_point(data = resampled_shags_dev, aes(color = bird_id)) +
  coord_sf(crs = '+proj=aeqd +lon_0=-68 +lat_0=-67 +units=m +datum=WGS84 +no_defs')+
  ggtitle("Resampled, old function")

patchwork::wrap_plots(plot_1, plot_2, plot_3, guides = "collect")
```

End.


