---
title: "Comparing kernels: adehabitatHR and tidytracks"
author: "Lizzie Pearmain"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
library(adehabitatHR)
library(MASS)
library(reshape2)
library(dplyr)
```

## genetta data

```{r}
genetta <- as.data.frame(matrix(
  c(
    0.5519758, 0.27524548,
    0.5725632, 0.12309273,
    0.5547747, 0.06100429,
    0.6110925, 0.16211416,
    0.5951087, 0.09316814,
    0.5333567, 0.11673812,
    0.5855461, 0.11170616,
    0.5221387, 0.11061583,
    0.5848452, 0.17213175
  ),
  ncol = 2, byrow = TRUE
))
```

adehabitatHR

```{r}
# this grid to be used in both
my_grid <- expand.grid(x = seq(0.01, 1, by = 0.01), 
                            y = seq(-0.4, 0.59, by = 0.01))

# kernel using adehabitat
mask_xy_grid <- my_grid
sp::coordinates(mask_xy_grid) <- ~ x + y
sp::gridded(mask_xy_grid) <- TRUE

g_ade <- adehabitatHR::kernelUD(sp::SpatialPoints(genetta),
  h = "href",
  grid = mask_xy_grid,
  kern = "bivnorm"
)

plot(g_ade)

```

MASS

```{r}
# kernel using MASS
mask.xy <- as.data.frame(my_grid)

H <- (sqrt(0.5 * (var(genetta[[1]]) + var(genetta[[2]])))) * (nrow(genetta)^-(1 / 6))

g_mass <- MASS::kde2d(genetta$V1,
  genetta$V2,
  n = c(100, 100), # TODO check this is right - it's still square
  h = c(H * 4, H * 4),
  lims = c(range(mask.xy$x), range(mask.xy$y))
)

# convert to spdf for plotting
g_mass_spdf <- sp::SpatialPixelsDataFrame(
  points = mask.xy,
  data = data.frame(est = reshape2::melt(g_mass$z)$value)
)
plot(g_mass_spdf)
```

Are they equal?

```{r}
# THIS is wrong, as it is VERY different
# likely flipped, possibly shifted
mean(abs(g_ade$ud - g_mass_spdf$est))
abs(mean(g_ade$ud - g_mass_spdf$est)) < 0.001

# flip it around
mean(abs(g_ade$ud - as.vector(matrix(g_mass$z,ncol=100, byrow=TRUE))))

# we are still slightly offset
which.max(g_ade$ud)
which.max(as.vector(matrix(g_mass$z,ncol=100, byrow=TRUE)))



```

```{r}
# clean up
rm(my_grid, mask_xy_grid, mask.xy, H)
rm(genetta, g_ade, g_mass, g_mass_spdf)
```


## puechabon example

```{r}
data(puechabonsp)

brock <- as.data.frame(puechabonsp$relocs) %>%
  dplyr::filter(Name == "Brock") %>%
  dplyr::select(x = X, y = Y)

# this grid to be used in both
my_grid <- expand.grid(x = seq(697000, 701950, by = 50), 
                       y = seq(3158550, 3163500, by = 50))

```


adehabitat

```{r}
# kernel using adehabitat
mask_xy_grid <- my_grid
sp::coordinates(mask_xy_grid) <- ~ x + y
sp::gridded(mask_xy_grid) <- TRUE

brock_ade <- adehabitatHR::kernelUD(sp::SpatialPoints(brock),
  h = "href",
  grid = mask_xy_grid,
  kern = "bivnorm"
)

plot(brock_ade)
```

MASS

```{r}
# kernel using MASS
mask.xy <- as.data.frame(my_grid)

H <- (sqrt(0.5 * (var(brock[[1]]) + var(brock[[2]])))) * (nrow(brock)^-(1 / 6))

brock_mass <- MASS::kde2d(brock$x,
  brock$y,
  n = c(100, 100), # TODO check this is right - it's still square
  h = c(H * 4, H * 4),
  lims = c(range(mask.xy$x), range(mask.xy$y))
)

# convert to spdf for plotting
brock_mass_spdf <- sp::SpatialPixelsDataFrame(
  points = mask.xy,
  data = data.frame(est = reshape2::melt(brock_mass$z)$value)
)
plot(brock_mass_spdf)
```

Are they equal?

```{r}

abs(mean(brock_ade$ud - brock_mass_spdf$est))
abs(mean(brock_ade$ud - brock_mass_spdf$est)) < 0.001

```

# Fit kernel with tidytracks

# load the boar data
```{r}
library(tidytracks)
wildboars_tt <- readRDS("./articles_data/wildboar_tt.rds")
wildboars_tt
```

Check the metadata
```{r}
show_meta(wildboars_tt)
```

Compute the kde with tidytracks. We need to be careful how we set the bounding box.
In terra, the extent is the corners, but kde works on centroids of cells.
TODO we need to check what ade habitat is doing, is the bbox same as a terra extent.
Are the individual coordinates the centroids (do we have individual coordinates?!?)

yes, it looks like the bbox of the grid is the same as the extent of the grid, so we can use the same values for both. We just need to make sure we are using the same resolution (we have 50 here, but we need to check what adehabitat is doing).

Note that the limits in the MASS kde function refer to centroids (i.e. the determine the actual values of x and y)
```{r}
wildboars_tt <- wildboars_tt %>% dplyr::group_by(Name)
wildboars_kde <- tt_hr_kde(wildboars_tt, levels = NULL,
                           h = "h_ref_indiv",
                           res = 50,
                           bbox = list(xmin = min(mask.xy$x)-25, ymin = min(mask.xy$y)-25,
                                    xmax = max(mask.xy$x)+25, ymax = max(mask.xy$y)+25))

```


check that we have the same h values (we do)
```{r}
brock_ade@h$h == wildboars_kde$h[1]

```

Check that the grid is the same
```{r}
all.equal(wildboars_kde$kde[[1]]$x, brock_mass$x)
all.equal(wildboars_kde$kde[[1]]$y, brock_mass$y)
all.equal(wildboars_kde$kde[[1]]$x, mask.xy$x[1:100])

```

```{r}
abs(mean(brock_ade$ud - wildboars_kde$kde[[1]]$z))
abs(mean(brock_ade$ud - wildboars_kde$kde[[1]]$z)) < 0.001
# this should be zero as we are comparing mass with mass 
# (just check that we are using the correct grid)
abs(mean(wildboars_kde$kde[[1]]$z - brock_mass_spdf$est))
```

Now transfer values from the tidytracks kernel to the adehabitatHR object, so
that we can check various operations

```{r}
brock_ade$ud <- as.vector(wildboars_kde$kde[[1]]$z)
# and confirm that now the values are identical
abs(mean(brock_ade$ud - wildboars_kde$kde[[1]]$z))
# plot it to check that it works
plot(brock_ade)
```

No create isopleths
```{r}
iso_90 <- getverticeshr(x = brock_ade,
                        percent = 90)
class(iso_90)
plot(iso_90, col=1:4)
```

```{r cumulative_ud}

vud <-  getvolumeUD(brock_ade)
vud
class(vud)

sum(vud@vol)

vud_std <- getvolumeUD(brock_ade, standardize = TRUE)
sum(vud_std@vol)

x <- brock_ade
        sg <- as(x, "SpatialPixelsDataFrame")
        sg <- as(sg, "SpatialGridDataFrame")
        gr <- gridparameters(sg)
        uu <- names(sg)
        gri <- as.image.SpatialGridDataFrame(sg)
  asc <- gri$z

just_std <- asc/(sum(asc))


```


## next steps

- try with `tidytracks::tt_hr_kde()` - getting the exact same grid etc. will be 
   difficult.
- check the differences are still <0.1%
- replace the `adehabitatHR` object's values (in `@data`) with the values from 
   the tidytracks kernel so they are exactly identical, just in different data
   formats from the different packages
- use this to test that the downstream functions work identically to the 
    `adehabitatHR` equivalents.
- we need to assign a projection to this dataset. I think it is 
  Lambert II Étendu (EPSG:27572).
  for the original dataset, it is something like:
proj4string(wildboar$relocations) <- CRS("+init=epsg:27572")





```{r}

```

















