---
title: "Comparing kernels: adehabitatHR and tidytracks"
author: "Lizzie Pearmain"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
library(adehabitatHR)
library(MASS)
library(reshape2)
library(dplyr)
```

## genetta data

```{r}
genetta <- as.data.frame(matrix(
  c(
    0.5519758, 0.27524548,
    0.5725632, 0.12309273,
    0.5547747, 0.06100429,
    0.6110925, 0.16211416,
    0.5951087, 0.09316814,
    0.5333567, 0.11673812,
    0.5855461, 0.11170616,
    0.5221387, 0.11061583,
    0.5848452, 0.17213175
  ),
  ncol = 2, byrow = TRUE
))
```

adehabitatHR

```{r}
# this grid to be used in both
my_grid <- expand.grid(x = seq(0.01, 1, by = 0.01), 
                            y = seq(-0.4, 0.59, by = 0.01))

# kernel using adehabitat
mask_xy_grid <- my_grid
sp::coordinates(mask_xy_grid) <- ~ x + y
sp::gridded(mask_xy_grid) <- TRUE

g_ade <- adehabitatHR::kernelUD(sp::SpatialPoints(genetta),
  h = "href",
  grid = mask_xy_grid,
  kern = "bivnorm"
)

plot(g_ade)

```

MASS

```{r}
# kernel using MASS
mask.xy <- as.data.frame(my_grid)

H <- (sqrt(0.5 * (var(genetta[[1]]) + var(genetta[[2]])))) * (nrow(genetta)^-(1 / 6))

g_mass <- MASS::kde2d(genetta$V1,
  genetta$V2,
  n = c(100, 100), # TODO check this is right - it's still square
  h = c(H * 4, H * 4),
  lims = c(range(mask.xy$x), range(mask.xy$y))
)

# convert to spdf for plotting
g_mass_spdf <- sp::SpatialPixelsDataFrame(
  points = mask.xy,
  data = data.frame(est = reshape2::melt(g_mass$z)$value)
)
plot(g_mass_spdf)
```

Are they equal?

```{r}

abs(mean(g_ade$ud - g_mass_spdf$est))
abs(mean(g_ade$ud - g_mass_spdf$est)) < 0.001

```

```{r}
# clean up
rm(my_grid, mask_xy_grid, mask.xy, H)
rm(genetta, g_ade, g_mass, g_mass_spdf)
```


## puechabon example

```{r}
data(puechabonsp)

brock <- as.data.frame(puechabonsp$relocs) %>%
  dplyr::filter(Name == "Brock") %>%
  dplyr::select(x = X, y = Y)

# this grid to be used in both
my_grid <- expand.grid(x = seq(697000, 701950, by = 50), 
                       y = seq(3158550, 3163500, by = 50))

```


adehabitat

```{r}
# kernel using adehabitat
mask_xy_grid <- my_grid
sp::coordinates(mask_xy_grid) <- ~ x + y
sp::gridded(mask_xy_grid) <- TRUE

brock_ade <- adehabitatHR::kernelUD(sp::SpatialPoints(brock),
  h = "href",
  grid = mask_xy_grid,
  kern = "bivnorm"
)

plot(brock_ade)
```

MASS

```{r}
# kernel using MASS
mask.xy <- as.data.frame(my_grid)

H <- (sqrt(0.5 * (var(brock[[1]]) + var(brock[[2]])))) * (nrow(brock)^-(1 / 6))

brock_mass <- MASS::kde2d(brock$x,
  brock$y,
  n = c(100, 100), # TODO check this is right - it's still square
  h = c(H * 4, H * 4),
  lims = c(range(mask.xy$x), range(mask.xy$y))
)

# convert to spdf for plotting
brock_mass_spdf <- sp::SpatialPixelsDataFrame(
  points = mask.xy,
  data = data.frame(est = reshape2::melt(brock_mass$z)$value)
)
plot(brock_mass_spdf)
```

Are they equal?

```{r}

abs(mean(brock_ade$ud - brock_mass_spdf$est))
abs(mean(brock_ade$ud - brock_mass_spdf$est)) < 0.001

```

# Fit kernel with tidytracks

# load the boar data
```{r}
library(tidytracks)
wildboars_tt <- readRDS("./articles_data/wildboar_mt.rds")
wildboars_tt
```

Check the metadata
```{r}
show_meta(wildboars_tt)
```

Compute the kde
```{r}
wildboars_tt <- wildboars_tt %>% dplyr::group_by(Name)
wildboars_kde <- tt_hr_kde(wildboars_tt, levels = NULL,
                           h = "h_ref_indiv",
                           res = 50,
                           bbox = list(xmin = brock_ade@bbox[1,1], ymin = brock_ade@bbox[2,1],
                                    xmax = brock_ade@bbox[1,2], ymax = brock_ade@bbox[2,2]))
```


check that we have the same h values (we do)
```{r}
brock_ade@h$h == wildboars_kde$h[1]

```

```{r}
abs(mean(brock_ade$ud - wildboars_kde$kde[[1]]$z))
abs(mean(brock_ade$ud - wildboars_kde$kde[[1]]$z)) < 0.001
```

## next steps

- try with `tidytracks::tt_hr_kde()` - getting the exact same grid etc. will be 
   difficult.
- check the differences are still <0.1%
- replace the `adehabitatHR` object's values (in `@data`) with the values from 
   the tidytracks kernel so they are exactly identical, just in different data
   formats from the different packages
- use this to test that the downstream functions work identically to the 
    `adehabitatHR` equivalents.





```{r}

```

















