---
title: "tidytracks alongside move2"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`tidytracks` alongside `move2`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Set up and load some packages and land for plotting
```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(tidytracks)
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
land <- ne_countries(scale = "medium", returnclass = "sf")
```

`tidytracks` is built on the `move2` object, and the two packages are designed 
to work together, though they have different philosophies and grammars. `move2` 
is automatically loaded when you load `tidytracks`, so you can use functions from 
both packages. Which package philosophy you prefer  to use is ultimately up to
you— the packages are fully inter-operable, and loading one won't mask the other.

We use the `move2` object because it's an excellent, tidyverse-friendly data 
structure that stores metadata memory-efficiently and carries it through the 
whole analysis alongside the event data (see our vignette for more detail on 
working with `move2` objects). The `move2` package uses this data structure and
provides data organisation tools and a limited set of specialised downstream 
analyses (e.g. those available in `move_2_utils`.

`tidytracks` builds on `move2`'s data structure to add extra functionality, with 
the goal of completing a full workflow within one package on one object. To do 
this, it uses a different grammar and function-naming convention. A coherent 
grammar, where function names reflect both their input and output, makes the 
package easier to learn and makes complex piped expressions more intuitive to 
write, since the order of functions follows naturally from their names.

The core of this grammar is that a `tidytracks` function's prefix tells you both
the object it operates on and the object it returns. For example, using the 
common `dplyr` verb `mutate()` to add a new column to a `move2` object requires 
a vector of the same length as the number of events (points in space and time) —
so it must operate on the output of an `event` function.

Many `tidytracks` functions mirror those in `move2`, often as simple wrappers 
that bring them into the `tidytracks` grammar; others have been reworked for 
distance. In contrast, `move2`, follows more traditional `sf` grammar and relies 
almost exclusively on `sf` under the hood. `tidytracks` departs from this, 
optimising such functions for speed, distance and home-range analysis 
(e.g. `event_distance()`). 

Because we use the `move2` data structure, events and home-range data are still 
returned as `sf` geometries, so are interoperable with `sf` functions, and come 
with a defined crs for easy plotting in `ggplot2`.

Below, we outline a simple workflow demonstrating the `tidytracks` grammar. 

First let's read in some example data: 

```{r read_events, message=FALSE}
shags_csv <- system.file(
  "extdata",
  "shags_example.csv",
  package = "tidytracks")

shags_tt <- tt_read_data(
  events = shags_csv,
  col_track_id = "bird_id",
  col_coords = c("lon", "lat"),
  col_date_time = "date_time",
  time_zone = "UTC",
  crs = 4326
)
```

Now that we have our data in a `move2` object, we can use `tidytracks` functions
to manipulate it. We can use a `tt` function which works on a whole `move2` 
object and returns another full `move2` table. For example, may want to ensure 
that the data is ordered by time within each track_id:
```{r, order time}
shags_tt_time <- shags_tt %>%
  tt_order_time()
```

Next, to use one of our computationally fast `event` functions, we can calculate 
the distance of each event:
```{r distance}
shags_distance <- shags_tt_time %>%
  event_distance()
```

As it is an `event` function, `event_distance()`  returns a vector of the same 
length as the numberof events (padded with an NA as distance is calculated
betweentwo points) which we can add to our `move2` object with `mutate()`:
```{r distance_mutate}
shags_tt_distance <- shags_tt_time %>%
  mutate(distance = shags_distance)
```

Because we know that the `tt` function `tt_order_time()` requires a whole `move2`
object, then as an `event` function, the output of `event_distance()` 
will be a vector, these steps could intuitively be implemented in one simple 
pipe:
```{r distance_mutate_pipe}
shags_tt_distance <- shags_tt %>%
  tt_order_time() %>%
  mutate(distance = event_distance(.))
```
Note that a `.` is required to pass the `move2` object to the `event_distance()` 
function. 

Lastly, with `sf` and `ggplot2` compatibility, it is easy to use functions 
such as our special `filter_by_meta()` alongside the `geom` plotting functions 
within `ggplot2` blocks to plot spatial data. Again, the grammar guides the user 
as to the intuitive order to add expressions into the pipe. 

```{r plot_track, fig.width=6, fig.height=6}
shags_tt_distance %>%
  mutate(
    speed = units::set_units(event_speed(.), "km/h")) %>%
  ggplot() +
    geom_sf(data = land, fill = "white") +
    geom_event_path(
      data = shags_tt_distance %>% filter_by_meta(bird_id == "kb_17"), 
      aes(color = speed), linewidth = 1.2) +
    coord_sf(xlim = c(-68.1, -67.65), ylim = c(-67.6, -67.2)) +
    scale_color_viridis_c()+
    labs(color = "Speed (km/hr)") +
    theme(aspect.ratio = 1)
  
```

