---
title: "tidytracks alongside move2"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`tidytracks` alongside `move2`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(tidytracks)
library(dplyr)
```


`tidytracks` is built on the `move2` object, and the two packages are designed to work together, though they have different philosophies and grammars.

We use the `move2` object because it's an excellent, tidyverse-friendly data structure that stores metadata memory-efficiently and carries it through the whole analysis alongside the event data (see our vignette for more detail on working with `move2` objects). The `move2` package uses this data structure and provides data organisation tools and a limited set of specialised downstream analyses.

`tidytracks` builds on `move2`'s data structure to add extra functionality, with the goal of completing a full workflow within one package on one object. To do this, it uses a different grammar and function-naming convention. A coherent grammar, where function names reflect both their input and output, makes the package easier to learn and makes complex piped expressions more intuitive to write, since the order of functions follows naturally from their names.

The core of this grammar is that a `tidytracks` function's prefix tells you both the object it operates on and the object it returns. For example, using the common `dplyr` verb `mutate()` to add a new column to a `move2` object requires a vector of the same length as the number of events (points in space and time) — so it must operate on the output of an `event` function.

Many `tidytracks` functions mirror those in `move2`, often as simple wrappers that bring them into the `tidytracks` grammar; others have been reworked for speed. In contrast, `move2`, follows more traditional `sf` grammar and relies almost exclusively on `sf` under the hood, including for calculations like speed and distance. `tidytracks` departs from this, optimising such functions for speed (e.g. `tt_event_speed()`).

Below, we outline a small workflow demonstrating the `tidytracks` grammar and its interoperability with `move2`. Which philosophy you prefer is ultimately up to you — the packages are fully interoperable, and loading one won't mask the other. ( TODO!)




1) 
```{r read_events, message=FALSE}
shags_csv <- system.file(
  "extdata",
  "shags_example.csv",
  package = "tidytracks")

shags_tt <- tt_read_data(
  events = shags_csv,
  col_track_id = "bird_id",
  col_coords = c("lon", "lat"),
  col_date_time = "date_time",
  time_zone = "UTC",
  crs = 4326
)
```

2) Now that we have our data in a `move2` object, we can use `tidytracks` functions
to manipulate it. For example, we can calculate the speed of each event:

```{r speed}
shags_tt_speed <- shags_tt %>%
  event_speed()
```

As it is an `event` function, it returns a vector of the same length as the number
of events (padded with an NA as speed is caluculated between two points) which we
can add to our `move2` object with `mutate()`:

```{r speed_mutate}
shags_tt <- shags_tt %>%
  mutate(speed = shags_tt_speed)
```
Because we know that the outpuit of `event_speed()` will be a vector, this could
also be done in one pipe. Note that a `.` is required to pass the `move2` object 
to the `event_speed()` function. 

```{r speed_mutate_pipe}
shags_tt <- shags_tt %>%
  mutate(speed = event_speed(.))
```


