---
title: "Untitled"
author: "HP"
date: "2026-06-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r setup, message=FALSE, warning=FALSE}
library(tidytracks)
library(sf)
library(dplyr)
```

## Load Example Dataset

```{r load-data}
data("shags_tt", package = "tidytracks")

shags_tt
```

## Inspect Columns and Metadata

```{r inspect}
colnames(shags_tt)

show_meta(shags_tt)
```

## Convert Metadata to Event Columns

Because `shags_tt` is a `move2` object with metadata attached, renaming the ID col breaks it

```{r metadata-to-events}
event_cols <- colnames(show_meta(shags_tt))

shags_tt_events <- shags_tt %>%
  as_event_column(all_of(event_cols))

show_meta(shags_tt_events)
```

## Convert to a DF so can change colname

```{r create-dataframe}
coords <- st_coordinates(shags_tt_events)

shags_df <- shags_tt_events %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  mutate(
    longitude = coords[, 1],
    latitude = coords[, 2]
  )

head(shags_df)
```

## Rename Track Identifier

```{r rename-id}
shags_df <- shags_df %>%
  rename(track_id = bird_id)

colnames(shags_df)
```

## make TT object again

```{r recreate-tidytracks}
shags <- tt_read_data(
  events = shags_df,
  col_track_id = "track_id",
  col_coords = c("longitude", "latitude"),
  col_date_time = "date_time",
  crs = 4326,
  time_zone = "UTC"
)

show_meta(shags)
```

## Inspect Track IDs

```{r inspect-track-ids}
unique(shags$track_id)
```

## Remove the track that violates the latitude filter.

```{r filter-track}
shags <- shags %>%
  filter(track_id != "KB_23_42962_38")
```

## Save 

```{r save-data}
save(shags, file = "data/shags_new.rda")
```

## check

```{r verify-save}
file.exists("data/shags_new.rda")
```

This can be pasted directly into an .Rmd file and knitted.